---
title: "Interactive Analysis: Volatility in Agricultural Commodities with MS-GARCH Models"
description: "A high-performance replication of the Master Thesis 'Modelling Volatility Using Markov-Switching GARCH Models' featuring interactive visualizations and deep econometric diagnostics."
author: "Rodrigo Hermont Ozon"
date: "2026-02-23"
categories: [Finance, R, MSGARCH, Commodities, Interactive, Risk Management]
format:
  html:
    toc: true
    toc-depth: 4
    code-fold: true
    code-summary: "Show the code"
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
---

# 1. Introduction

Modelling financial time-series has for decades been subject to extensive research. In the Master thesis *"Modelling Volatility Using Markov-Switching GARCH Models"* (Gluud & Jacobsen, 2021), the authors thoroughly examined whether the introduction of **Markov-Switching (MS) regimes** to the classic Generalized Autoregressive Conditional Heteroscedasticity (GARCH) framework improves in-sample modelling and out-of-sample risk forecasting for the S&P 500 index. 

This post replicates and expands that research, applying it to an **Equally Weighted Portfolio of Agricultural Commodities**. We leverage **interactive visualizations** to better understand the volatility dynamics and the transition between market regimes.

---

# 2. Data Preparation and Stylized Facts

```{r}
#| label: setup-and-data

library(quantmod)
library(MSGARCH)
library(PerformanceAnalytics)
library(plotly)
library(tidyr)
library(knitr)
library(kableExtra)
library(moments)
library(dplyr)

# Fetch data
tickers <- c("ZC=F", "ZO=F", "KE=F", "ZR=F", "GF=F", "ZS=F", "ZM=F", "ZL=F")
getSymbols(tickers, from = "2015-01-01", to = "2023-12-31", warnings = FALSE)

prices <- do.call(merge, lapply(tickers, function(x) Ad(get(x))))
prices <- na.locf(prices)
returns <- Return.calculate(prices, method = "log")[-1]

# Equally weighted portfolio
port_returns <- rowMeans(returns, na.rm = TRUE) * 100
df_plot <- data.frame(Date = index(returns), Returns = as.numeric(port_returns))
ret_vec <- as.numeric(port_returns)
```

## 2.1. Interactive Volatility Clustering

Volatility clustering is visually evident. Use the range slider below to zoom into specific market events.

```{r}
#| label: eda-returns-plotly
#| fig-cap: "Interactive Portfolio Returns and Squared Returns"

p1 <- plot_ly(df_plot, x = ~Date, y = ~Returns, type = 'scatter', mode = 'lines', 
              name = 'Returns (%)', line = list(color = '#2c3e50')) %>%
  layout(yaxis = list(title = "Returns (%)"), xaxis = list(rangeslider = list(visible = T)))

p2 <- plot_ly(df_plot, x = ~Date, y = ~Returns^2, type = 'scatter', mode = 'lines', 
              name = 'Sq. Returns', line = list(color = '#c0392b')) %>%
  layout(yaxis = list(title = "Squared Returns"))

subplot(p1, p2, nrows = 2, shareX = TRUE, titleY = TRUE) %>%
  layout(showlegend = FALSE, title = "Agricultural Commodities Portfolio: Clustering Visualization")
```

## 2.2. Non-Normality: Histogram & Density

The commodities portfolio exhibits significant **leptokurtosis** (kurtosis > 3).

```{r}
#| label: eda-distribution-plotly
#| fig-cap: "Distribution Analysis: Empirical vs Normal"

# Prepare density data
dens <- density(ret_vec)
norm_dens <- dnorm(dens$x, mean = mean(ret_vec), sd = sd(ret_vec))

fig_dist <- plot_ly() %>%
  add_histogram(x = ~ret_vec, name = "Histogram", marker = list(color = '#bdc3c7', line = list(color = 'white', width = 1))) %>%
  add_lines(x = dens$x, y = dens$y, name = "Empirical Density", line = list(color = '#2c3e50', width = 3), yaxis = "y2") %>%
  add_lines(x = dens$x, y = norm_dens, name = "Normal Density", line = list(color = '#c0392b', width = 2, dash = 'dash'), yaxis = "y2") %>%
  layout(
    title = "Portfolio Return Distribution",
    xaxis = list(title = "Returns (%)"),
    yaxis = list(title = "Frequency"),
    yaxis2 = list(overlaying = "y", side = "right", title = "Density", showgrid = FALSE),
    legend = list(orientation = 'h', y = -0.2)
  )

fig_dist

cat("Skewness:", round(skewness(ret_vec), 2), "
")
cat("Kurtosis:", round(kurtosis(ret_vec), 2), "
")
```

## 2.3. Autocorrelation (ACF) Diagnostics

```{r}
#| label: eda-acf-plotly
#| fig-cap: "Interactive ACF Plots"

calc_acf_df <- function(x, title) {
  a <- acf(x, plot = FALSE, lag.max = 20)
  df <- data.frame(Lag = a$lag, ACF = a$acf)
  conf <- qnorm((1 + 0.95)/2)/sqrt(a$n.used)
  return(list(df = df, conf = conf, title = title))
}

acf_ret <- calc_acf_df(ret_vec, "Returns ACF")
acf_sq <- calc_acf_df(ret_vec^2, "Squared Returns ACF")

plot_acf <- function(obj) {
  plot_ly(obj$df, x = ~Lag, y = ~ACF, type = 'bar', name = obj$title, marker = list(color = '#34495e')) %>%
    add_segments(x = 0, xend = 20, y = obj$conf, yend = obj$conf, line = list(color = 'red', dash = 'dash'), name = '95% CI') %>%
    add_segments(x = 0, xend = 20, y = -obj$conf, yend = -obj$conf, line = list(color = 'red', dash = 'dash'), showlegend = FALSE) %>%
    layout(yaxis = list(range = c(-0.2, 1)), title = obj$title)
}

subplot(plot_acf(acf_ret), plot_acf(acf_sq), nrows = 1, titleX = TRUE)
```

---

# 3. Model Estimation

```{r}
#| label: model-fitting
#| results: hide

spec_gn <- CreateSpec(variance.spec = list(model = "sGARCH"), distribution.spec = list(distribution = "norm"))
spec_gs <- CreateSpec(variance.spec = list(model = "sGARCH"), distribution.spec = list(distribution = "std"))
spec_mn <- CreateSpec(variance.spec = list(model = c("sGARCH", "sGARCH")), distribution.spec = list(distribution = c("norm", "norm")))
spec_ms <- CreateSpec(variance.spec = list(model = c("sGARCH", "sGARCH")), distribution.spec = list(distribution = c("std", "std")))

fit_gn <- FitML(spec = spec_gn, data = ret_vec)
fit_gs <- FitML(spec = spec_gs, data = ret_vec)
fit_mn <- FitML(spec = spec_mn, data = ret_vec)
fit_ms <- FitML(spec = spec_ms, data = ret_vec)
```

---

# 4. Results & Diagnostics

## 4.1. Information Criteria

```{r}
#| label: in-sample-comparison

results <- data.frame(
  Model = c("GARCH-Norm", "GARCH-Std", "MS-GARCH-Norm", "MS-GARCH-Std"),
  LogLik = c(fit_gn$loglik, fit_gs$loglik, fit_mn$loglik, fit_ms$loglik),
  AIC = c(AIC(fit_gn), AIC(fit_gs), AIC(fit_mn), AIC(fit_ms)),
  BIC = c(BIC(fit_gn), BIC(fit_gs), BIC(fit_mn), BIC(fit_ms))
)

results |> kable(digits = 2) |> kable_styling()
```

## 4.2. Latent Regimes & Conditional Volatility

```{r}
#| label: smoothed-prob-plotly
#| fig-cap: "Dynamic Volatility and Regime Probabilities (MS-tGARCH)"

sp <- State(fit_ms)
df_ms <- data.frame(
  Date = index(returns),
  Prob_High_Vol = sp$SmoothProb[, 1, 1][-1],
  Volatility = Volatility(fit_ms)
)

fig_regime <- plot_ly(df_ms, x = ~Date) %>%
  add_lines(y = ~Volatility, name = "Volatility", line = list(color = '#2980b9')) %>%
  add_lines(y = ~Prob_High_Vol, name = "Prob. Regime 1 (High)", line = list(color = '#c0392b', width = 1), fill = 'tozeroy', yaxis = "y2", opacity = 0.3) %>%
  layout(
    title = "Conditional Volatility vs. Regime Probability",
    yaxis = list(title = "Volatility"),
    yaxis2 = list(overlaying = "y", side = "right", title = "Probability", range = c(0, 1)),
    xaxis = list(rangeslider = list(visible = T))
  )

fig_regime
```

## 4.3. Standardized Residual Analysis

```{r}
#| label: residual-plotly
#| fig-cap: "Interactive QQ-Plot Comparison"

res_gn <- ret_vec / Volatility(fit_gn)
res_gs <- ret_vec / Volatility(fit_gs)

qq_data <- function(res, name) {
  n <- length(res)
  df <- data.frame(
    Theoretical = qnorm(ppoints(n)),
    Sample = sort(res),
    Model = name
  )
  return(df)
}

df_qq <- rbind(qq_data(res_gn, "GARCH-Norm"), qq_data(res_gs, "GARCH-Std"))

plot_ly(df_qq, x = ~Theoretical, y = ~Sample, color = ~Model, colors = c('#7f8c8d', '#2c3e50'), type = 'scatter', mode = 'markers', marker = list(size = 4, opacity = 0.6)) %>%
  add_segments(x = -4, xend = 4, y = -4, yend = 4, line = list(color = 'red', width = 1), showlegend = FALSE) %>%
  layout(title = "QQ-Plot: Normal vs Student-t Residuals", xaxis = list(title = "Theoretical Quantiles"), yaxis = list(title = "Sample Quantiles"))
```

---

# 5. Risk Assessment (VaR & Expected Shortfall)

```{r}
#| label: risk-evaluation-results

risk_gn <- Risk(fit_gn, alpha = 0.05, do.es = TRUE, do.its = TRUE)
risk_gs <- Risk(fit_gs, alpha = 0.05, do.es = TRUE, do.its = TRUE)
risk_mn <- Risk(fit_mn, alpha = 0.05, do.es = TRUE, do.its = TRUE)
risk_ms <- Risk(fit_ms, alpha = 0.05, do.es = TRUE, do.its = TRUE)

calc_v <- function(risk_obj, rets) {
  var_viol <- sum(rets < -risk_obj$VaR)
  es_viol <- sum(rets < -risk_obj$ES)
  return(c(var_viol, es_viol))
}

risk_matrix <- rbind(calc_v(risk_gn, ret_vec), calc_v(risk_gs, ret_vec), calc_v(risk_mn, ret_vec), calc_v(risk_ms, ret_vec))
rownames(risk_matrix) <- c("GARCH-Norm", "GARCH-Std", "MS-GARCH-Norm", "MS-GARCH-Std")
colnames(risk_matrix) <- c("VaR Violations", "ES Violations")

risk_matrix |> kable() |> kable_styling()
```

---

# 6. Conclusion

By using **interactive Plotly visualizations**, we can see precisely where GARCH models with normal distributions fail to capture the "jumps" in agricultural markets. The **MS-GARCH-Std** model provides the most comprehensive framework for capturing both structural shifts and extreme tail risks.

***
*Interactive Report generated via Quarto and Plotly.*
