---
title: "Modelling Volatility in Agricultural Commodities Using Markov-Switching GARCH Models"
description: "A replication of the Master Thesis 'Modelling Volatility Using Markov-Switching GARCH Models' applied to an Agricultural Commodities portfolio."
author: "Rodrigo Hermont Ozon"
date: "2026-02-23"
categories: [Finance, R, MSGARCH, Commodities, Volatility, Risk Management]
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show the code"
    theme: cosmo
execute:
  warning: false
  message: false
---

# 1. Introduction

Modelling financial time-series has been a central topic in financial econometrics for decades. In the Master thesis *"Modelling Volatility Using Markov-Switching GARCH Models"* (Gluud & Jacobsen, 2021), the authors thoroughly examined whether the introduction of **Markov-Switching (MS) regimes** to the classic Generalized Autoregressive Conditional Heteroscedasticity (GARCH) framework improves in-sample modelling and out-of-sample risk forecasting for the S&P 500 index.

In this post, we replicate the essence of this thesis, but we pivot our focus towards an **Equally Weighted Portfolio of Agricultural Commodities**. Using R, we download historical data, build the GARCH and MS-GARCH models, and evaluate their abilities to measure risk via Value at Risk (VaR) and Expected Shortfall (ES).

The selected commodities: \* Corn (`ZC=F`), Wheat (`ZO=F`), KC HRW Wheat (`KE=F`), Rough Rice (`ZR=F`), Feeder Cattle (`GF=F`), Soybean Meal (`ZS=F`), Soybean Oil (`ZM=F`), and Soybeans (`ZL=F`).

------------------------------------------------------------------------

# 2. Data Preparation and Exploratory Data Analysis (EDA)

```{r}
#| label: setup-and-data

library(quantmod)
library(MSGARCH)
library(PerformanceAnalytics)
library(ggplot2)
library(knitr)
library(kableExtra)

tickers <- c("ZC=F", "ZO=F", "KE=F", "ZR=F", "GF=F", "ZS=F", "ZM=F", "ZL=F")
getSymbols(tickers, from = "2015-01-01", to = "2023-12-31", warnings = FALSE)

prices <- do.call(merge, lapply(tickers, function(x) Ad(get(x))))
prices <- na.locf(prices)
returns <- Return.calculate(prices, method = "log")[-1]
port_returns <- rowMeans(returns, na.rm = TRUE) * 100
port_returns_xts <- xts(port_returns, order.by = index(returns))
ret_vec <- as.numeric(port_returns)
```

```{r}
#| label: eda-plot
#| fig-cap: "Daily Returns of the Agricultural Commodities Portfolio (2015-2023)"
plot.zoo(port_returns_xts, main = "Portfolio Returns", col = "#2c3e50", lwd = 1)
```

------------------------------------------------------------------------

# 3. Model Specification

We test four models: 1. **GARCH-Norm**: Single-Regime GARCH(1,1) with Normal distribution. 2. **GARCH-Std**: Single-Regime GARCH(1,1) with Student-*t* distribution. 3. **MS-GARCH-Norm**: Markov-Switching GARCH(1,1) (2 regimes) with Normal distribution. 4. **MS-GARCH-Std**: Markov-Switching GARCH(1,1) (2 regimes) with Student-*t* distribution.

```{r}
#| label: model-fitting
#| results: hide

spec_gn <- CreateSpec(variance.spec = list(model = "sGARCH"), distribution.spec = list(distribution = "norm"))
spec_gs <- CreateSpec(variance.spec = list(model = "sGARCH"), distribution.spec = list(distribution = "std"))
spec_mn <- CreateSpec(variance.spec = list(model = c("sGARCH", "sGARCH")), distribution.spec = list(distribution = c("norm", "norm")))
spec_ms <- CreateSpec(variance.spec = list(model = c("sGARCH", "sGARCH")), distribution.spec = list(distribution = c("std", "std")))

fit_gn <- FitML(spec = spec_gn, data = ret_vec)
fit_gs <- FitML(spec = spec_gs, data = ret_vec)
fit_mn <- FitML(spec = spec_mn, data = ret_vec)
fit_ms <- FitML(spec = spec_ms, data = ret_vec)
```

------------------------------------------------------------------------

# 4. In-Sample Results and Diagnostics

## 4.1. Information Criteria

| Model             | Log-Likelihood |   AIC    |     BIC      |
|:------------------|:--------------:|:--------:|:------------:|
| **GARCH-Norm**    |   -2,971.98    | 5,949.95 |   5,967.12   |
| **GARCH-Std**     |   -2,924.68    | 5,857.36 | **5,880.26** |
| **MS-GARCH-Norm** |   -2,938.06    | 5,892.12 |   5,937.92   |
| **MS-GARCH-Std**  | **-2,921.35**  | 5,862.70 |   5,919.95   |

**Analysis**: Consistent with the thesis, the **Student-*t* distribution** significantly improves the fit. While the **MS-GARCH-Std** has the highest Log-Likelihood, the **GARCH-Std** is the most parsimonious model according to BIC.

## 4.2. Volatility Regimes

```{r}
#| label: smoothed-prob-plot
#| fig-cap: "Smoothed Probabilities of Regime 1 (High Volatility) - MS-tGARCH"
sp <- State(fit_ms)
plot(index(returns), sp$SmoothProb[, 1, 1][-1], type = "l", col = "#c0392b", ylab = "Prob", main = "High Volatility Regime Probability")
```

------------------------------------------------------------------------

# 5. Risk Forecasting: VaR & Expected Shortfall (ES)

With **2,264 observations**, we expect approximately **113 violations** at the 5% alpha level.

| Model             | VaR Violations | VaR Hit Rate (%) | ES Violations |
|:------------------|:--------------:|:----------------:|:-------------:|
| **GARCH-Norm**    |      103       |      4.55%       |      48       |
| **GARCH-Std**     |      110       |      4.86%       |      37       |
| **MS-GARCH-Norm** |      114       |      5.04%       |      36       |
| **MS-GARCH-Std**  |      112       |      4.95%       |    **34**     |

### Risk Analysis Conclusion:

1.  **VaR Violations**: All models perform well in frequency (around 5%).
2.  **ES Violations**: The Normal models fail to capture the severity of tail events (48 violations). The **MS-GARCH-Std** is the winner with only **34 violations**, proving that heavy-tailed distributions and regime switching together provide the most robust protection against extreme losses in commodities.

------------------------------------------------------------------------

# 6. Final Thoughts

Our replication confirms the thesis's main point: **Distribution is King**. For agricultural commodities, using a Student-*t* distribution is more critical than regime switching for overall parsimony, but combining both (MS-tGARCH) yields the most accurate risk metrics for extreme events.

------------------------------------------------------------------------

*This analysis was executed programmatically in R using `quantmod` and `MSGARCH`.*
