---
title: "Commodity News Monitor"
subtitle: "Weekly Monitoring, PAIC Commodities Portfolio"
date: today
date-format: "DD/MM/YYYY"
author: "Rodrigo Hermont Ozon - FAE"
format:
  html:
    theme: cosmo
    toc: true
    toc-title: "Navigation"
    toc-location: left
    toc-depth: 2
    number-sections: false
    self-contained: true
    code-fold: true
    code-tools: true
    code-summary: "Show Code"
    df-print: paged
    css: news_styles.css
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
pkgs <- c("tidyRSS", "httr2", "dplyr", "purrr", "glue",
          "lubridate", "stringr", "readr", "jsonlite",
          "htmltools", "tidyr", "quantmod", "blastula", "ggplot2", "scales")

invisible(lapply(pkgs, function(p) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p)
  library(p, character.only = TRUE)
}))

# quantmod/xts/zoo mascaram dplyr — forcamos de volta
filter <- dplyr::filter
select <- dplyr::select

DAYS_BACK   <- 7
SLEEP_SEC   <- 1.5
OUTPUT_DIR  <- "data"
NEWSAPI_KEY <- Sys.getenv("NEWSAPI_KEY")
GROQ_KEY    <- Sys.getenv("GROQ_API_KEY")
GROQ_MODEL  <- "llama-3.3-70b-versatile"
SMTP_USER   <- Sys.getenv("SMTP_USER")
SMTP_PASS   <- Sys.getenv("SMTP_PASS")
EMAIL_TO    <- Sys.getenv("EMAIL_TO")

# MC parameters
N_SIM   <- 5000   # simulacoes Monte Carlo
T_DAYS  <- 30     # horizonte em dias uteis

# Pesos dos criterios fuzzy (somam 1)
FUZZY_WEIGHTS <- c(
  supply_disruption     = 0.25,
  demand_momentum       = 0.25,
  geopolitical_risk     = 0.15,
  speculative_sentiment = 0.20,
  macro_headwind        = 0.15
)

# Direcao: +1 = bullish, -1 = bearish
FUZZY_DIRECTION <- c(
  supply_disruption     = -1,
  demand_momentum       =  1,
  geopolitical_risk     = -1,
  speculative_sentiment =  1,
  macro_headwind        = -1
)

if (!dir.exists(OUTPUT_DIR)) dir.create(OUTPUT_DIR, recursive = TRUE)

ticker_map <- tibble::tribble(
  ~ticker,  ~nome,            ~query_google,                         ~query_newsapi,
  "ZC=F",   "Corn",           "corn futures commodities market",      "corn futures",
  "ZO=F",   "Oat",            "oat grain futures market price",       "oat futures grain",
  "KE=F",   "HRW Wheat",      "wheat futures Kansas HRW grain",       "wheat futures grain market",
  "ZR=F",   "Rough Rice",     "rice futures commodity market price",  "rice futures commodity",
  "GF=F",   "Feeder Cattle",  "cattle futures livestock market",      "cattle futures livestock",
  "ZS=F",   "Soybeans",       "soybean futures commodities market",   "soybean futures",
  "ZM=F",   "Soybean Meal",   "soybean meal futures market",          "soybean meal futures",
  "ZL=F",   "Soybean Oil",    "soybean oil futures market",           "soybean oil futures"
)

news_blacklist <- list(
  global  = c("bitcoin", "crypto", "ethereum", "nft", "blockchain", "dogecoin",
              "poker", "casino", "lottery", "horoscope"),
  "ZC=F"  = c("popcorn", "candy corn", "corn maze", "corn dog"),
  "ZO=F"  = c("oat milk brand", "oatly"),
  "GF=F"  = c("feeder fund", "venture", "startup", "hedge fund"),
  "ZR=F"  = c("rough cut", "rough draft", "rough rider")
)

empty_news <- function() {
  tibble(
    ticker = character(), nome = character(), fonte = character(),
    titulo = character(), snippet = character(), link = character(),
    veiculo = character(), publicado = as.POSIXct(character())
  )
}
```

```{r fetch_prices}
make_sparkline_svg <- function(prices, width = 80, height = 28) {
  prices <- as.numeric(prices[!is.na(prices)])
  if (length(prices) < 2) return("")
  mn <- min(prices); mx <- max(prices); rng <- ifelse(mx == mn, 1, mx - mn)
  xs  <- round((seq_along(prices) - 1) / (length(prices) - 1) * width, 2)
  ys  <- round((1 - (prices - mn) / rng) * height, 2)
  pts <- paste(xs, ys, sep = ",", collapse = " ")
  color <- if (tail(prices, 1) >= prices[1]) "#16a34a" else "#dc2626"
  glue('<svg width="{width}" height="{height}" viewBox="0 0 {width} {height}" ',
       'xmlns="http://www.w3.org/2000/svg" style="display:inline-block;vertical-align:middle;">',
       '<polyline points="{pts}" fill="none" stroke="{color}" stroke-width="1.5" ',
       'stroke-linejoin="round" stroke-linecap="round"/></svg>')
}

fetch_prices <- function(tickers, days_back = 90) {
  purrr::map_dfr(tickers, function(tk) {
    tryCatch({
      raw <- getSymbols(tk, src = "yahoo", auto.assign = FALSE,
                        from = Sys.Date() - days_back - 5, to = Sys.Date())
      cls <- as.numeric(Cl(raw)); cls <- cls[!is.na(cls)]; n <- length(cls)
      if (n < 5) return(tibble(ticker=tk, price=NA_real_, weekly_chg=NA_real_,
                               sparkline="", mu_daily=NA_real_, sigma_daily=NA_real_))
      price      <- tail(cls, 1)
      price_1w   <- if (n >= 6) cls[n - 5] else cls[1]
      weekly_chg <- round((price / price_1w - 1) * 100, 2)
      # GBM parameters: log-returns diarios
      log_rets    <- diff(log(cls))
      mu_daily    <- mean(log_rets, na.rm = TRUE)
      sigma_daily <- sd(log_rets,   na.rm = TRUE)
      tibble(ticker = tk, price = round(price, 2),
             weekly_chg = weekly_chg,
             sparkline  = make_sparkline_svg(tail(cls, 30)),
             mu_daily   = mu_daily,
             sigma_daily = sigma_daily)
    }, error = function(e)
      tibble(ticker=tk, price=NA_real_, weekly_chg=NA_real_,
             sparkline="", mu_daily=NA_real_, sigma_daily=NA_real_))
  })
}

prices_df <- fetch_prices(ticker_map$ticker)
```

```{r filter_functions}
filter_news <- function(df, ticker_key) {
  if (nrow(df) == 0) return(df)
  words <- c(news_blacklist[["global"]], news_blacklist[[ticker_key]] %||% character())
  if (length(words) == 0) return(df)
  pattern <- paste(words, collapse = "|")
  dplyr::filter(df, !grepl(pattern, tolower(titulo), fixed = FALSE))
}
```

```{r fetch_functions}
fetch_google_news <- function(ticker, nome, query_google, days_back = DAYS_BACK, ...) {
  url <- glue("https://news.google.com/rss/search?q={URLencode(query_google)}&hl=en-US&gl=US&ceid=US:en")
  tryCatch({
    feed <- tidyRSS::tidyfeed(url, clean_tags = TRUE)
    if (is.null(feed) || nrow(feed) == 0) return(empty_news())
    veiculo_col <- if ("item_feed_title" %in% names(feed)) feed$item_feed_title
                   else if ("feed_title"  %in% names(feed)) feed$feed_title
                   else rep("Google News", nrow(feed))
    snippet_col <- if ("item_description" %in% names(feed))
                     stringr::str_trunc(feed$item_description, 300)
                   else rep(NA_character_, nrow(feed))
    result <- feed |>
      dplyr::mutate(
        ticker = ticker, nome = nome, fonte = "Google News",
        publicado = with_tz(item_pub_date, "America/Sao_Paulo"),
        veiculo = veiculo_col, snippet = snippet_col
      ) |>
      dplyr::filter(publicado >= Sys.time() - days(days_back)) |>
      dplyr::select(ticker, nome, fonte, titulo = item_title, snippet,
                    link = item_link, veiculo, publicado)
    if (nrow(result) == 0) empty_news() else result
  }, error = function(e) {
    message(glue("[Google] Error {ticker}: {conditionMessage(e)}")); empty_news()
  })
}

fetch_newsapi <- function(ticker, nome, query_newsapi, days_back = DAYS_BACK, ...) {
  if (is.na(NEWSAPI_KEY) || NEWSAPI_KEY == "") return(empty_news())
  tryCatch({
    resp <- request("https://newsapi.org/v2/everything") |>
      req_url_query(q = query_newsapi, from = format(Sys.Date() - days_back, "%Y-%m-%d"),
                    sortBy = "publishedAt", language = "en",
                    pageSize = 10, apiKey = NEWSAPI_KEY) |>
      req_perform() |> resp_body_json(simplifyVector = TRUE)
    arts <- resp$articles
    if (is.null(arts) || length(arts) == 0) return(empty_news())
    arts <- as_tibble(arts)
    if (nrow(arts) == 0) return(empty_news())
    result <- arts |>
      dplyr::mutate(
        ticker = ticker, nome = nome, fonte = "NewsAPI",
        publicado = as.POSIXct(publishedAt, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC") |>
                    with_tz("America/Sao_Paulo"),
        veiculo = if (is.data.frame(source)) source$name else as.character(source),
        snippet = stringr::str_trunc(description %||% NA_character_, 300)
      ) |>
      dplyr::filter(publicado >= Sys.time() - days(days_back)) |>
      dplyr::select(ticker, nome, fonte, titulo = title, snippet,
                    link = url, veiculo, publicado)
    if (nrow(result) == 0) empty_news() else result
  }, error = function(e) {
    message(glue("[NewsAPI] Error {ticker}: {conditionMessage(e)}")); empty_news()
  })
}

fetch_all_news <- function(days_back = DAYS_BACK) {
  ticker_map |>
    purrr::pmap_dfr(function(ticker, nome, query_google, query_newsapi) {
      message(glue("  Fetching: {nome} ({ticker})"))
      g <- fetch_google_news(ticker, nome, query_google, days_back)
      Sys.sleep(SLEEP_SEC)
      n <- fetch_newsapi(ticker, nome, query_newsapi, days_back)
      combined <- bind_rows(g, n)
      if (nrow(combined) > 0 && any(!is.na(combined$titulo) & combined$titulo != "")) {
        combined <- combined |>
          dplyr::filter(!is.na(titulo), titulo != "") |>
          dplyr::distinct(titulo, .keep_all = TRUE) |>
          filter_news(ticker_key = ticker) |>
          dplyr::arrange(desc(publicado))
      }
      combined
    })
}
```

```{r load_news}
news_df <- fetch_all_news()
readr::write_csv(news_df, file.path(OUTPUT_DIR, glue("commodity_news_{Sys.Date()}.csv")))

total_news <- nrow(dplyr::filter(news_df, !is.na(titulo), titulo != ""))
period     <- glue("{format(Sys.Date() - DAYS_BACK, '%m/%d')} to {format(Sys.Date(), '%m/%d/%Y')}")

ticker_counts <- ticker_map |>
  dplyr::left_join(
    dplyr::filter(news_df, !is.na(titulo), titulo != "") |> dplyr::count(ticker, name = "n"),
    by = "ticker"
  ) |>
  dplyr::mutate(n = ifelse(is.na(n), 0L, n))

headlines_by_ticker <- news_df |>
  dplyr::filter(!is.na(titulo), titulo != "") |>
  dplyr::group_by(ticker, nome) |>
  dplyr::summarise(
    headlines_rich = paste(
      ifelse(!is.na(snippet) & snippet != "",
             paste0(titulo, "\n  > ", snippet), titulo),
      collapse = "\n\n"
    ),
    n_articles = dplyr::n(),
    .groups = "drop"
  )
```

```{r groq_functions}
groq_summarize <- function(headlines_text, commodity_name, model = GROQ_MODEL) {
  if (is.na(GROQ_KEY) || GROQ_KEY == "") return(NULL)
  prompt <- paste0(
    "You are a commodity market analyst. Below are news headlines and article snippets ",
    "from the past week about ", commodity_name, " futures markets.\n\n",
    "Headlines & Snippets:\n", headlines_text, "\n\n",
    "Write a concise market summary (3-5 sentences) covering:\n",
    "1. The main price drivers or trends mentioned\n",
    "2. Key supply/demand factors\n",
    "3. Any relevant geopolitical or weather events\n",
    "4. Overall market sentiment (bullish/bearish/neutral)\n\n",
    "Use details from the snippets when available. Be direct and analytical. ",
    "Do not repeat headlines verbatim."
  )
  tryCatch({
    resp <- request("https://api.groq.com/openai/v1/chat/completions") |>
      req_headers("Authorization" = paste("Bearer", GROQ_KEY), "Content-Type" = "application/json") |>
      req_body_json(list(model = model,
        messages = list(
          list(role = "system", content = "You are a concise commodity market analyst. Always respond in English."),
          list(role = "user",   content = prompt)
        ), temperature = 0.3, max_tokens = 350)) |>
      req_timeout(30) |> req_perform() |> resp_body_json()
    resp$choices[[1]]$message$content
  }, error = function(e) NULL)
}

generate_all_summaries <- function(df_rich) {
  if (is.na(GROQ_KEY) || GROQ_KEY == "")
    return(tibble(ticker = character(), nome = character(), summary = character()))
  df_rich |>
    dplyr::mutate(summary = purrr::map2_chr(headlines_rich, nome, function(h, name) {
      Sys.sleep(0.5)
      result <- groq_summarize(h, name)
      if (is.null(result)) "No news available this week." else result
    })) |>
    dplyr::select(ticker, nome, summary)
}

summaries_df <- generate_all_summaries(headlines_by_ticker)

summaries_df <- ticker_map |>
  dplyr::select(ticker, nome) |>
  dplyr::left_join(summaries_df, by = c("ticker", "nome")) |>
  dplyr::mutate(summary = ifelse(is.na(summary),
                                 "No news found for this commodity this week.", summary))

readr::write_csv(summaries_df, file.path(OUTPUT_DIR, glue("summaries_{Sys.Date()}.csv")))
```

```{r groq_portfolio}
groq_portfolio_summary <- function(summaries_df) {
  if (is.na(GROQ_KEY) || GROQ_KEY == "") return(NULL)
  all_summaries <- summaries_df |>
    dplyr::mutate(block = glue("{nome} ({ticker}): {summary}")) |>
    dplyr::pull(block) |> paste(collapse = "\n\n")
  prompt <- paste0(
    "You are a senior commodity portfolio analyst. Below are individual weekly summaries ",
    "for each commodity in an agricultural futures portfolio.\n\n", all_summaries, "\n\n",
    "Write a comprehensive portfolio-level market brief (5-7 sentences) that:\n",
    "1. Identifies the dominant macro theme across the portfolio this week\n",
    "2. Highlights the top 2-3 commodities with the most significant moves or news\n",
    "3. Notes any cross-commodity correlations or common drivers (weather, USD, energy)\n",
    "4. Gives an overall portfolio sentiment (risk-on / risk-off / mixed)\n",
    "5. Flags any key events or reports to watch next week\n\n",
    "Be direct, analytical, and concise. Write as a professional market brief."
  )
  tryCatch({
    resp <- request("https://api.groq.com/openai/v1/chat/completions") |>
      req_headers("Authorization" = paste("Bearer", GROQ_KEY), "Content-Type" = "application/json") |>
      req_body_json(list(model = GROQ_MODEL,
        messages = list(
          list(role = "system", content = "You are a senior commodity portfolio analyst. Always respond in English."),
          list(role = "user",   content = prompt)
        ), temperature = 0.3, max_tokens = 500)) |>
      req_timeout(40) |> req_perform() |> resp_body_json()
    resp$choices[[1]]$message$content
  }, error = function(e) NULL)
}

groq_price_outlook <- function(ticker, nome, headlines_text) {
  if (is.na(GROQ_KEY) || GROQ_KEY == "" || nchar(headlines_text) < 10) return(NULL)
  prompt <- paste0(
    "Based on these news headlines and snippets about ", nome, " (", ticker, ") futures:\n\n",
    headlines_text, "\n\n",
    "Return ONLY a JSON object (no markdown) with exactly these fields:\n",
    '{"horizon_days": <7 or 30 or 90>, "expectation": "<bullish|bearish|neutral>", ',
    '"confidence": "<high|medium|low>", "key_factor": "<one sentence max>"}'
  )
  tryCatch({
    resp <- request("https://api.groq.com/openai/v1/chat/completions") |>
      req_headers("Authorization" = paste("Bearer", GROQ_KEY), "Content-Type" = "application/json") |>
      req_body_json(list(model = GROQ_MODEL,
        messages = list(
          list(role = "system", content = "Return only valid JSON. No markdown, no backticks."),
          list(role = "user",   content = prompt)
        ), temperature = 0.1, max_tokens = 150)) |>
      req_timeout(20) |> req_perform() |> resp_body_json()
    raw <- gsub("```json|```", "", resp$choices[[1]]$message$content) |> trimws()
    jsonlite::fromJSON(raw)
  }, error = function(e) NULL)
}

portfolio_summary <- groq_portfolio_summary(summaries_df)

outlook_list <- headlines_by_ticker |>
  dplyr::mutate(outlook = purrr::map2(ticker, nome, function(tk, nm) {
    h <- headlines_by_ticker$headlines_rich[headlines_by_ticker$ticker == tk]
    Sys.sleep(0.3)
    groq_price_outlook(tk, nm, h)
  }))
```

```{r fuzzy_mc}
# ═══════════════════════════════════════════════════════════════════════════════
# FUZZY MULTI-CRITERIA + MONTE CARLO (GBM)
# Pipeline:
#   1. Groq extrai scores fuzzy [0,1] para 5 criterios por ticker (via LLM)
#   2. Agregacao multicritério: net_signal ∈ [-1,1] via soma ponderada direcional
#   3. Calibracao GBM: μ e σ diarios estimados dos retornos historicos (quantmod)
#   4. Ajuste de drift: μ_adj = μ_hist + net_signal * σ * k
#   5. Monte Carlo GBM: S(T) = S0 * exp((μ_adj - σ²/2)*T + σ*√T*Z), Z~N(0,1)
#   6. Output: E[ret], VaR(5%), Prob(ret>0), intervalo [P5, P95]
# ═══════════════════════════════════════════════════════════════════════════════

# ── 1. Extracao fuzzy via Groq ─────────────────────────────────────────────────
groq_fuzzy_extraction <- function(ticker, nome, headlines_rich) {
  if (is.na(GROQ_KEY) || GROQ_KEY == "" || nchar(headlines_rich) < 10) {
    # Sem noticias: retorna scores neutros (0.5)
    return(tibble(ticker = ticker, nome = nome,
                  supply_disruption = 0.5, demand_momentum = 0.5,
                  geopolitical_risk = 0.5, speculative_sentiment = 0.5,
                  macro_headwind = 0.5, source = "neutral_default"))
  }
  prompt <- paste0(
    "You are extracting structured market signals from commodity news.\n\n",
    "News about ", nome, " (", ticker, ") futures:\n\n", headlines_rich, "\n\n",
    "Score each criterion from 0.0 to 1.0 based solely on evidence in the news:\n",
    "- supply_disruption: 0=no disruption, 1=severe supply disruption (bearish)\n",
    "- demand_momentum: 0=weak/falling demand, 1=strong/rising demand (bullish)\n",
    "- geopolitical_risk: 0=no geopolitical tension, 1=high geopolitical risk (bearish)\n",
    "- speculative_sentiment: 0=bearish market positioning, 1=bullish positioning\n",
    "- macro_headwind: 0=no macro headwinds, 1=strong headwinds (strong USD, high rates) (bearish)\n\n",
    "Use 0.5 for any criterion with insufficient evidence in the news.\n",
    "Return ONLY valid JSON:\n",
    '{"supply_disruption":X,"demand_momentum":X,"geopolitical_risk":X,',
    '"speculative_sentiment":X,"macro_headwind":X}'
  )
  tryCatch({
    resp <- request("https://api.groq.com/openai/v1/chat/completions") |>
      req_headers("Authorization" = paste("Bearer", GROQ_KEY), "Content-Type" = "application/json") |>
      req_body_json(list(model = GROQ_MODEL,
        messages = list(
          list(role = "system", content = "Return only valid JSON with numeric values 0.0-1.0. No markdown."),
          list(role = "user",   content = prompt)
        ), temperature = 0.1, max_tokens = 120)) |>
      req_timeout(20) |> req_perform() |> resp_body_json()
    raw    <- gsub("```json|```", "", resp$choices[[1]]$message$content) |> trimws()
    parsed <- jsonlite::fromJSON(raw)
    # Clipa todos os scores para [0, 1]
    parsed <- lapply(parsed, function(x) max(0, min(1, as.numeric(x))))
    tibble(ticker = ticker, nome = nome,
           supply_disruption     = parsed$supply_disruption     %||% 0.5,
           demand_momentum       = parsed$demand_momentum       %||% 0.5,
           geopolitical_risk     = parsed$geopolitical_risk     %||% 0.5,
           speculative_sentiment = parsed$speculative_sentiment %||% 0.5,
           macro_headwind        = parsed$macro_headwind        %||% 0.5,
           source = "groq_extracted")
  }, error = function(e) {
    message(glue("[Fuzzy] Error {ticker}: {conditionMessage(e)}"))
    tibble(ticker = ticker, nome = nome,
           supply_disruption = 0.5, demand_momentum = 0.5,
           geopolitical_risk = 0.5, speculative_sentiment = 0.5,
           macro_headwind = 0.5, source = "error_default")
  })
}

# Extrai criterios para todos os tickers (com e sem noticias)
all_tickers_headlines <- ticker_map |>
  dplyr::left_join(headlines_by_ticker |> dplyr::select(ticker, headlines_rich),
                   by = "ticker") |>
  dplyr::mutate(headlines_rich = ifelse(is.na(headlines_rich), "", headlines_rich))

fuzzy_scores <- all_tickers_headlines |>
  purrr::pmap_dfr(function(ticker, nome, headlines_rich, ...) {
    Sys.sleep(0.4)
    groq_fuzzy_extraction(ticker, nome, headlines_rich)
  })

# ── 2. Agregacao multicritério ─────────────────────────────────────────────────
criteria_cols <- names(FUZZY_WEIGHTS)

fuzzy_scores <- fuzzy_scores |>
  dplyr::rowwise() |>
  dplyr::mutate(
    # net_signal ∈ [-1, 1]: soma ponderada de (direcao * (score - 0.5) * 2)
    # score=0.5 → contribuicao zero (neutro)
    # score=1.0 + direcao=+1 → contribuicao = +peso (max bullish)
    # score=0.0 + direcao=-1 → contribuicao = +peso (max bearish efeito no sinal negativo)
    net_signal = sum(purrr::map_dbl(criteria_cols, function(c) {
      FUZZY_DIRECTION[c] * FUZZY_WEIGHTS[c] * (get(c) - 0.5) * 2
    })),
    scenario = dplyr::case_when(
      net_signal >  0.15 ~ "Optimistic",
      net_signal < -0.15 ~ "Pessimistic",
      TRUE               ~ "Base"
    )
  ) |>
  dplyr::ungroup()

# ── 3 & 4. GBM + Ajuste de drift + Monte Carlo ────────────────────────────────
set.seed(42)  # reproducibilidade

simulate_gbm <- function(S0, mu_daily, sigma_daily, net_signal,
                          T = T_DAYS, n = N_SIM) {
  if (is.na(S0) || is.na(mu_daily) || is.na(sigma_daily))
    return(NULL)
  # Ajuste conservador: net_signal escala o drift em ate 50% de 1 sigma
  k          <- 0.5
  mu_adj     <- mu_daily + net_signal * sigma_daily * k
  # GBM closed-form: S(T) = S0 * exp((mu - sigma²/2)*T + sigma*sqrt(T)*Z)
  Z          <- rnorm(n)
  log_ret    <- (mu_adj - 0.5 * sigma_daily^2) * T + sigma_daily * sqrt(T) * Z
  pct_ret    <- (exp(log_ret) - 1) * 100
  list(
    pct_returns   = pct_ret,
    E_ret         = mean(pct_ret),
    VaR_5         = quantile(pct_ret, 0.05),
    VaR_1         = quantile(pct_ret, 0.01),
    prob_pos      = mean(pct_ret > 0) * 100,
    P5            = quantile(pct_ret, 0.05),
    P25           = quantile(pct_ret, 0.25),
    P75           = quantile(pct_ret, 0.75),
    P95           = quantile(pct_ret, 0.95),
    mu_adj        = mu_adj,
    sigma_daily   = sigma_daily,
    net_signal    = net_signal
  )
}

# Junta tudo
mc_inputs <- fuzzy_scores |>
  dplyr::left_join(prices_df |> dplyr::select(ticker, price, mu_daily, sigma_daily),
                   by = "ticker")

mc_results <- mc_inputs |>
  dplyr::mutate(mc = purrr::pmap(
    list(price, mu_daily, sigma_daily, net_signal),
    function(p, mu, sig, ns) simulate_gbm(p, mu, sig, ns)
  ))

# Salva resultados
mc_summary <- mc_results |>
  dplyr::mutate(
    E_ret    = purrr::map_dbl(mc, ~ if (is.null(.x)) NA_real_ else .x$E_ret),
    VaR_5    = purrr::map_dbl(mc, ~ if (is.null(.x)) NA_real_ else .x$VaR_5),
    prob_pos = purrr::map_dbl(mc, ~ if (is.null(.x)) NA_real_ else .x$prob_pos),
    P5       = purrr::map_dbl(mc, ~ if (is.null(.x)) NA_real_ else .x$P5),
    P95      = purrr::map_dbl(mc, ~ if (is.null(.x)) NA_real_ else .x$P95)
  ) |>
  dplyr::select(ticker, nome, scenario, net_signal, E_ret, VaR_5, prob_pos, P5, P95,
                supply_disruption, demand_momentum, geopolitical_risk,
                speculative_sentiment, macro_headwind)

readr::write_csv(mc_summary |> dplyr::select(-c(supply_disruption:macro_headwind)),
                 file.path(OUTPUT_DIR, glue("mc_scenario_{Sys.Date()}.csv")))
```

---

::: {.callout-note}
**Period covered:** `r period` &nbsp;|&nbsp; **Total articles:** `r total_news` &nbsp;|&nbsp; **Sources:** Google News RSS + NewsAPI.org &nbsp;|&nbsp; **AI:** Groq / llama-3.3-70b
:::

---

## Portfolio Overview

```{r portfolio_overview}
price_table <- ticker_map |>
  dplyr::left_join(prices_df, by = "ticker") |>
  dplyr::left_join(dplyr::select(ticker_counts, ticker, n), by = "ticker")

rows <- price_table |>
  purrr::pmap(function(ticker, nome, price, weekly_chg, sparkline, n, ...) {
    chg_color <- if (is.na(weekly_chg)) "#6b7280" else if (weekly_chg >= 0) "#16a34a" else "#dc2626"
    chg_arrow <- if (is.na(weekly_chg)) "" else if (weekly_chg >= 0) "\u25b2" else "\u25bc"
    chg_txt   <- if (is.na(weekly_chg)) "N/A" else paste0(chg_arrow, " ", abs(weekly_chg), "%")
    price_txt <- if (is.na(price)) "N/A" else formatC(price, format = "f", digits = 2, big.mark = ",")
    max_n     <- max(price_table$n, 1)
    bar_w     <- round((n / max_n) * 60)
    bar_color <- if (n == 0) "#e5e7eb" else if (n >= 5) "#2c5f2e" else "#86b48a"
    tags$tr(
      tags$td(style = "font-weight:600; color:#1a1a2e; padding:.45rem .6rem;", nome),
      tags$td(style = "padding:.45rem .6rem;",
              tags$code(style = "font-size:.75rem; color:#6b7280;", ticker)),
      tags$td(style = "font-family:monospace; text-align:right; padding:.45rem .6rem;", price_txt),
      tags$td(style = glue("color:{chg_color}; font-weight:600; text-align:right; padding:.45rem .6rem;"), chg_txt),
      tags$td(style = "padding:.45rem .6rem;", HTML(sparkline)),
      tags$td(style = "padding:.45rem .6rem;",
        div(style = "display:flex; align-items:center; gap:.4rem;",
          div(style = glue("height:10px; width:{bar_w}px; min-width:3px; background:{bar_color}; border-radius:3px;")),
          tags$span(style = "font-size:.75rem; color:#9ca3af;", n)
        )
      )
    )
  })

div(class = "portfolio-panel",
  tags$table(style = "width:100%; border-collapse:collapse;",
    tags$thead(tags$tr(
      lapply(c("Commodity","Ticker","Price","7d Change","30d Chart","Articles"),
             function(h) tags$th(style = "text-align:left; padding:.4rem .6rem; font-size:.78rem;
                                          color:#6b7280; border-bottom:2px solid #e5e7eb;
                                          font-weight:600; text-transform:uppercase;", h))
    )),
    tags$tbody(lapply(rows, function(r)
      htmltools::tagAppendAttributes(r, style = "border-bottom:1px solid #f3f4f6;")))
  )
)
```

## Portfolio AI Brief

```{r portfolio_brief}
if (is.null(portfolio_summary) || is.na(GROQ_KEY) || GROQ_KEY == "") {
  div(class = "callout callout-warning",
      p("Portfolio AI brief unavailable. Configure GROQ_API_KEY to enable."))
} else {
  outlook_rows <- outlook_list |>
    dplyr::filter(!purrr::map_lgl(outlook, is.null)) |>
    purrr::pmap(function(ticker, nome, outlook, ...) {
      exp   <- outlook$expectation  %||% "neutral"
      conf  <- outlook$confidence   %||% "low"
      horiz <- outlook$horizon_days %||% 7
      key   <- outlook$key_factor   %||% ""
      exp_color  <- switch(exp,  bullish="#16a34a", bearish="#dc2626", "#d97706")
      conf_color <- switch(conf, high="#1e40af", medium="#7c3aed", "#6b7280")
      arrow      <- switch(exp,  bullish="\u25b2", bearish="\u25bc", "\u25cf")
      exp_bg     <- switch(exp,  bullish="#f0fdf4", bearish="#fef2f2", "#fffbeb")
      tags$tr(
        tags$td(style="padding:5px 10px;font-weight:600;color:#1a1a2e;", nome),
        tags$td(style="padding:5px 10px;",
                tags$code(style="font-size:.72rem;color:#6b7280;", ticker)),
        tags$td(style="padding:5px 10px;text-align:center;color:#6b7280;", paste0(horiz,"d")),
        tags$td(style=glue("padding:5px 10px;text-align:center;background:{exp_bg};border-radius:4px;"),
          tags$span(style=glue("color:{exp_color};font-weight:700;font-size:.85rem;"), paste(arrow,exp))),
        tags$td(style=glue("padding:5px 10px;text-align:center;color:{conf_color};font-size:.8rem;font-weight:600;"), conf),
        tags$td(style="padding:5px 10px;font-size:.82rem;color:#4b5563;", key)
      )
    })

  outlook_table <- if (length(outlook_rows) > 0)
    div(style="margin-top:1.4rem;border-top:1px solid #dbeafe;padding-top:1rem;",
      tags$p(style="font-size:.75rem;font-weight:700;color:#6b7280;text-transform:uppercase;
                    letter-spacing:.05em;margin-bottom:.6rem;", "\U0001f4ca Price Outlook by Commodity"),
      tags$table(style="width:100%;border-collapse:collapse;font-size:.84rem;",
        tags$thead(tags$tr(style="border-bottom:2px solid #e5e7eb;",
          lapply(c("Commodity","Ticker","Horizon","Expectation","Confidence","Key Factor"),
                 function(h) tags$th(style="padding:4px 10px;text-align:left;color:#9ca3af;
                                            font-size:.73rem;font-weight:600;text-transform:uppercase;", h))
        )),
        tags$tbody(lapply(outlook_rows, function(r)
          htmltools::tagAppendAttributes(r, style="border-bottom:1px solid #f3f4f6;")))
      )
    ) else NULL

  div(class="portfolio-brief",
    div(class="brief-header",
      tags$span(class="brief-icon","AI"),
      tags$span(class="brief-title","Weekly Portfolio Brief"),
      tags$span(class="brief-model", glue("Groq / {GROQ_MODEL}"))
    ),
    div(class="brief-body", portfolio_summary),
    outlook_table
  )
}
```

---

## Weekly AI Summary

```{r weekly_overview}
sentiment_badge <- function(summary_text) {
  kw_bull <- c("rally","surge","gain","rise","high","strong","bull","upward","demand","support")
  kw_bear <- c("drop","fall","decline","weak","low","bear","downward","surplus","pressure","loss")
  s  <- tolower(summary_text)
  b  <- sum(sapply(kw_bull, grepl, x = s))
  br <- sum(sapply(kw_bear, grepl, x = s))
  if (b > br)      tags$span(class="sentiment bullish","Bullish")
  else if (br > b) tags$span(class="sentiment bearish","Bearish")
  else             tags$span(class="sentiment neutral", "Neutral")
}

cards <- summaries_df |>
  dplyr::left_join(dplyr::select(ticker_counts, ticker, n), by = "ticker") |>
  purrr::pmap(function(ticker, nome, summary, n, ...) {
    has_news <- n > 0
    div(class="summary-card",
      tags$h3(
        style=paste0("margin:0 0 .6rem 0;padding-bottom:.5rem;",
                     "border-bottom:2px solid ", if(has_news)"#2c5f2e" else "#e5e7eb",";",
                     "display:flex;align-items:center;gap:.6rem;flex-wrap:wrap;"),
        tags$span(style="font-size:1.1rem;font-weight:800;color:#1a1a2e;letter-spacing:-.01em;", nome),
        tags$code(style="font-size:.72rem;color:#6b7280;font-weight:400;background:#f3f4f6;
                         padding:2px 6px;border-radius:4px;", ticker),
        sentiment_badge(summary),
        tags$span(style="margin-left:auto;font-size:.78rem;color:#9ca3af;font-weight:400;",
                  paste(n,"articles"))
      ),
      div(class="summary-body",
          style=if(!has_news)"color:#9ca3af;font-style:italic;" else "",
          summary)
    )
  })

tagList(cards)
```

---

## News by Commodity

```{r render_cards}
make_news_card <- function(titulo, link, veiculo, publicado, fonte, snippet, ...) {
  badge_color <- if (!is.na(fonte) && fonte == "Google News") "#4285F4" else "#E84393"
  div(class="news-card",
    div(class="news-meta",
      tags$span(class="badge", style=paste0("background:",badge_color), fonte),
      tags$span(class="veiculo", veiculo),
      tags$span(class="data", format(publicado, "%m/%d %H:%M"))
    ),
    a(href=link, target="_blank", class="news-titulo", titulo),
    if (!is.na(snippet) && nchar(trimws(snippet)) > 0)
      tags$p(style="font-size:.78rem;color:#6b7280;margin:.3rem 0 0 0;line-height:1.4;",
             stringr::str_trunc(snippet, 180))
    else NULL
  )
}

render_ticker_section <- function(ticker_key, df_ticker) {
  nome_display <- unique(df_ticker$nome)[1]
  df_valid     <- dplyr::filter(df_ticker, !is.na(titulo), titulo != "")
  n_articles   <- nrow(df_valid)
  cards <- if (n_articles == 0)
    tags$blockquote(style="color:#9ca3af;border-left:3px solid #e5e7eb;padding-left:1rem;",
                    "No articles found for this ticker in the selected period.")
  else
    df_valid |> dplyr::arrange(desc(publicado)) |> purrr::pmap(make_news_card) |> tagList()
  tagList(
    tags$h2(tags$span(nome_display),
            tags$code(style="margin-left:.5rem;font-size:.75em;color:#6b7280;", ticker_key)),
    tags$p(tags$em(paste0(n_articles," article(s) found in the last ",DAYS_BACK," days"))),
    cards
  )
}

sections <- ticker_map |>
  purrr::pmap(function(ticker, nome, ...) {
    df_ticker <- dplyr::filter(news_df, ticker == !!ticker)
    if (nrow(df_ticker) == 0) {
      df_ticker <- empty_news() |>
        tibble::add_row(ticker=ticker, nome=nome, fonte=NA_character_,
                        titulo=NA_character_, snippet=NA_character_,
                        link=NA_character_, veiculo=NA_character_,
                        publicado=as.POSIXct(NA))
    }
    render_ticker_section(ticker, df_ticker)
  })

tagList(sections)
```

---

## Fuzzy-MC Scenario Analysis

::: {.callout-tip}
**Methodology:** News-driven fuzzy multi-criteria scoring → drift-adjusted GBM → Monte Carlo simulation.
For each commodity, Groq extracts 5 market signals from headlines+snippets (scores ∈ [0,1]).
A weighted directional aggregation produces a **net signal** ∈ [-1,1] that adjusts the historical GBM drift.
**N = `r N_SIM` simulations | Horizon = `r T_DAYS` trading days (~6 weeks)**

*Note: Short-term news have limited predictive power on futures prices — this framework is best interpreted as a structured qualitative scenario tool, not a forecast.*
:::

### Fuzzy Criteria Scores

```{r fuzzy_heatmap}
criteria_labels <- c(
  supply_disruption     = "Supply Disruption",
  demand_momentum       = "Demand Momentum",
  geopolitical_risk     = "Geopolitical Risk",
  speculative_sentiment = "Speculative Sentiment",
  macro_headwind        = "Macro Headwind"
)

score_color <- function(score, direction) {
  # direction +1: verde = alto, vermelho = baixo
  # direction -1: vermelho = alto, verde = baixo
  if (is.na(score)) return("background:#f9fafb; color:#9ca3af;")
  intensity <- abs(score - 0.5) * 2  # 0 a 1
  if (direction == 1) {
    if (score > 0.5) glue("background:rgba(22,163,74,{0.15 + intensity*0.35}); color:#14532d;")
    else             glue("background:rgba(220,38,38,{0.15 + intensity*0.35}); color:#7f1d1d;")
  } else {
    if (score > 0.5) glue("background:rgba(220,38,38,{0.15 + intensity*0.35}); color:#7f1d1d;")
    else             glue("background:rgba(22,163,74,{0.15 + intensity*0.35}); color:#14532d;")
  }
}

scenario_badge <- function(scenario) {
  switch(scenario,
    Optimistic  = tags$span(style="background:#dcfce7;color:#15803d;padding:2px 8px;
                                   border-radius:4px;font-size:.75rem;font-weight:700;", "\u25b2 Optimistic"),
    Pessimistic = tags$span(style="background:#fee2e2;color:#b91c1c;padding:2px 8px;
                                   border-radius:4px;font-size:.75rem;font-weight:700;", "\u25bc Pessimistic"),
    tags$span(style="background:#fef9c3;color:#854d0e;padding:2px 8px;
                     border-radius:4px;font-size:.75rem;font-weight:700;", "\u25cf Base")
  )
}

header_row <- tags$tr(
  tags$th(style="padding:6px 10px;text-align:left;color:#6b7280;font-size:.75rem;
                 font-weight:600;text-transform:uppercase;border-bottom:2px solid #e5e7eb;",
          "Commodity"),
  lapply(names(criteria_labels), function(c)
    tags$th(style="padding:6px 10px;text-align:center;color:#6b7280;font-size:.72rem;
                   font-weight:600;text-transform:uppercase;border-bottom:2px solid #e5e7eb;",
            criteria_labels[c])
  ),
  tags$th(style="padding:6px 10px;text-align:center;color:#6b7280;font-size:.75rem;
                 font-weight:600;text-transform:uppercase;border-bottom:2px solid #e5e7eb;",
          "Net Signal"),
  tags$th(style="padding:6px 10px;text-align:center;color:#6b7280;font-size:.75rem;
                 font-weight:600;text-transform:uppercase;border-bottom:2px solid #e5e7eb;",
          "Scenario")
)

data_rows <- fuzzy_scores |>
  purrr::pmap(function(ticker, nome, supply_disruption, demand_momentum,
                        geopolitical_risk, speculative_sentiment, macro_headwind,
                        net_signal, scenario, ...) {
    scores_vals <- c(supply_disruption, demand_momentum, geopolitical_risk,
                     speculative_sentiment, macro_headwind)
    score_cells <- purrr::map2(scores_vals, names(FUZZY_DIRECTION), function(s, c) {
      tags$td(style=paste0("padding:6px 10px;text-align:center;font-size:.82rem;font-weight:600;",
                           score_color(s, FUZZY_DIRECTION[c])),
              sprintf("%.2f", s))
    })
    net_color <- if (net_signal > 0.15) "color:#16a34a;font-weight:700;"
                 else if (net_signal < -0.15) "color:#dc2626;font-weight:700;"
                 else "color:#d97706;font-weight:700;"
    htmltools::tagAppendAttributes(
      tags$tr(
        tags$td(style="padding:6px 10px;font-weight:600;color:#1a1a2e;", nome),
        tagList(score_cells),
        tags$td(style=paste0("padding:6px 10px;text-align:center;font-size:.85rem;", net_color),
                sprintf("%+.3f", net_signal)),
        tags$td(style="padding:6px 10px;text-align:center;", scenario_badge(scenario))
      ),
      style = "border-bottom:1px solid #f3f4f6;"
    )
  })

div(class="portfolio-panel",
  tags$table(style="width:100%;border-collapse:collapse;font-size:.85rem;",
    tags$thead(header_row),
    tags$tbody(data_rows)
  )
)
```

### Monte Carlo Results

```{r mc_results_table}
mc_table_rows <- mc_results |>
  dplyr::filter(!purrr::map_lgl(mc, is.null)) |>
  purrr::pmap(function(ticker, nome, scenario, mc, price, ...) {
    e_ret    <- sprintf("%+.2f%%", mc$E_ret)
    var5     <- sprintf("%.2f%%", mc$VaR_5)
    p_pos    <- sprintf("%.1f%%", mc$prob_pos)
    range    <- glue("[{sprintf('%+.1f%%', mc$P5)} , {sprintf('%+.1f%%', mc$P95)}]")
    e_color  <- if (mc$E_ret > 0) "color:#16a34a;font-weight:700;"
                else if (mc$E_ret < 0) "color:#dc2626;font-weight:700;"
                else "color:#d97706;"
    pp_color <- if (mc$prob_pos > 55) "color:#16a34a;"
                else if (mc$prob_pos < 45) "color:#dc2626;"
                else "color:#d97706;"
    htmltools::tagAppendAttributes(
      tags$tr(
        tags$td(style="padding:6px 10px;font-weight:600;color:#1a1a2e;", nome),
        tags$td(style="padding:6px 10px;", tags$code(style="font-size:.72rem;color:#6b7280;", ticker)),
        tags$td(style="padding:6px 10px;text-align:center;", scenario_badge(scenario)),
        tags$td(style=paste0("padding:6px 10px;text-align:right;font-family:monospace;", e_color), e_ret),
        tags$td(style="padding:6px 10px;text-align:right;font-family:monospace;color:#dc2626;", var5),
        tags$td(style=paste0("padding:6px 10px;text-align:right;", pp_color), p_pos),
        tags$td(style="padding:6px 10px;text-align:center;font-family:monospace;font-size:.78rem;color:#6b7280;", range)
      ),
      style = "border-bottom:1px solid #f3f4f6;"
    )
  })

div(class="portfolio-panel",
  tags$p(style="font-size:.75rem;color:#6b7280;margin-bottom:.8rem;",
    glue("GBM calibrated on 90-day historical prices | Drift adjusted by fuzzy net signal | ",
         "N={scales::comma(N_SIM)} simulations | T={T_DAYS} trading days")),
  tags$table(style="width:100%;border-collapse:collapse;font-size:.85rem;",
    tags$thead(tags$tr(style="border-bottom:2px solid #e5e7eb;",
      lapply(c("Commodity","Ticker","Scenario","E[Return]","VaR (5%)","P(ret>0)","90% Range [P5,P95]"),
             function(h) tags$th(style="padding:6px 10px;text-align:left;color:#6b7280;
                                        font-size:.73rem;font-weight:600;text-transform:uppercase;", h))
    )),
    tags$tbody(mc_table_rows)
  )
)
```

### Return Distribution

```{r mc_density_plot, fig.width=10, fig.height=5}
# Monta dataframe com todas as distribuicoes simuladas
dist_data <- mc_results |>
  dplyr::filter(!purrr::map_lgl(mc, is.null)) |>
  dplyr::select(ticker, nome, scenario, mc) |>
  dplyr::mutate(returns = purrr::map(mc, ~ .x$pct_returns)) |>
  tidyr::unnest(returns)

scenario_colors <- c(Optimistic = "#16a34a", Base = "#d97706", Pessimistic = "#dc2626")

# Linhas verticais de VaR
var_lines <- mc_results |>
  dplyr::filter(!purrr::map_lgl(mc, is.null)) |>
  dplyr::mutate(VaR_5 = purrr::map_dbl(mc, ~ .x$VaR_5),
                E_ret  = purrr::map_dbl(mc, ~ .x$E_ret))

ggplot(dist_data, aes(x = returns, fill = scenario, color = scenario)) +
  geom_density(alpha = 0.25, linewidth = 0.6) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#6b7280", linewidth = 0.4) +
  geom_vline(data = var_lines, aes(xintercept = VaR_5, color = scenario),
             linetype = "dotted", linewidth = 0.8) +
  facet_wrap(~ nome, nrow = 2, scales = "free") +
  scale_fill_manual(values = scenario_colors) +
  scale_color_manual(values = scenario_colors) +
  scale_x_continuous(labels = scales::percent_format(scale = 1, accuracy = 1)) +
  labs(
    title    = glue("Monte Carlo Return Distribution — {T_DAYS}-Day Horizon"),
    subtitle = glue("N = {scales::comma(N_SIM)} simulations per commodity | GBM with news-adjusted drift"),
    x        = "Simulated Return (%)",
    y        = "Density",
    fill     = "Scenario",
    color    = "Scenario",
    caption  = "Dotted vertical lines = VaR(5%) | Dashed = zero return"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "#6b7280", size = 9),
    strip.text    = element_text(face = "bold", size = 9),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

---

```{r send_email}
send_weekly_email <- function() {
  if (any(c(SMTP_USER, SMTP_PASS, EMAIL_TO) == "")) {
    message("[Email] Credenciais nao configuradas — pulando."); return(invisible(NULL))
  }
  if (is.null(portfolio_summary)) {
    message("[Email] Portfolio summary indisponivel — pulando."); return(invisible(NULL))
  }
  price_rows_html <- price_table |>
    purrr::pmap_chr(function(ticker, nome, price, weekly_chg, n, ...) {
      chg_color <- if(is.na(weekly_chg))"#6b7280" else if(weekly_chg>=0)"#16a34a" else "#dc2626"
      chg_arrow <- if(is.na(weekly_chg))"" else if(weekly_chg>=0)"\u25b2" else "\u25bc"
      chg_txt   <- if(is.na(weekly_chg))"N/A" else paste0(chg_arrow,abs(weekly_chg),"%")
      price_txt <- if(is.na(price))"N/A" else formatC(price,format="f",digits=2,big.mark=",")
      glue('<tr>
        <td style="padding:6px 10px;border-bottom:1px solid #f3f4f6;font-weight:600;">{nome}</td>
        <td style="padding:6px 10px;border-bottom:1px solid #f3f4f6;color:#6b7280;">{ticker}</td>
        <td style="padding:6px 10px;border-bottom:1px solid #f3f4f6;text-align:right;">{price_txt}</td>
        <td style="padding:6px 10px;border-bottom:1px solid #f3f4f6;color:{chg_color};font-weight:600;text-align:right;">{chg_txt}</td>
        <td style="padding:6px 10px;border-bottom:1px solid #f3f4f6;text-align:center;color:#6b7280;">{n} art.</td>
      </tr>')
    }) |> paste(collapse="\n")

  email <- compose_email(body = blocks(
    block_title("Commodity News Monitor"),
    block_text(glue("Weekly brief — {period}")),
    block_spacer(),
    block_text("<strong>Portfolio Brief (AI)</strong>"),
    block_text(portfolio_summary),
    block_spacer(),
    block_text("<strong>Price Summary</strong>"),
    block_text(glue(
      '<table style="width:100%;border-collapse:collapse;font-size:.9rem;">
        <thead><tr style="background:#f9fafb;">
          <th style="padding:6px 10px;text-align:left;color:#6b7280;">Commodity</th>
          <th style="padding:6px 10px;text-align:left;color:#6b7280;">Ticker</th>
          <th style="padding:6px 10px;text-align:right;color:#6b7280;">Price</th>
          <th style="padding:6px 10px;text-align:right;color:#6b7280;">7d</th>
          <th style="padding:6px 10px;text-align:center;color:#6b7280;">News</th>
        </tr></thead><tbody>{price_rows_html}</tbody></table>'
    )),
    block_spacer(),
    block_text(glue(
      '<p style="font-size:.8rem;color:#9ca3af;">Auto-generated every Friday | PAIC Econometrics / FAE<br>',
      'Full report: <a href="https://paiceconometrics.github.io/site/news.html">View online</a></p>'
    ))
  ))

  tryCatch({
    smtp_send(email,
      to = strsplit(EMAIL_TO,",")[[1]] |> trimws(),
      from = SMTP_USER,
      subject = glue("Commodity Weekly Brief — {format(Sys.Date(),'%b %d, %Y')}"),
      credentials = creds_envvar(user="SMTP_USER", pass_envvar="SMTP_PASS",
                                 host="smtp.gmail.com", port=465, use_ssl=TRUE))
    message(glue("[Email] Enviado para: {EMAIL_TO}"))
  }, error = function(e) message(glue("[Email] Erro: {conditionMessage(e)}")))
}

send_weekly_email()
```

---

*Auto-report scheduled every Friday at 07:00 AM BRT | PAIC Econometrics / FAE Business School*
