---
title: "Research Methodology"
subtitle: "Systematic Approach to Multi-Objective Portfolio Optimization"
format:
  html:
    toc: true
    toc-depth: 4
    code-fold: true
    code-tools: true
---

# Research Methodology

## Overview

This research project adopts a **quantitative, computational, and applied** approach, combining statistical econometrics, machine learning, and optimization techniques to develop an integrated framework for agricultural commodities portfolio management.

Our methodology follows a rigorous quasi-experimental design based on time series simulations and backtesting protocols, ensuring reproducibility and transparency throughout all research phases.

::: {.callout-note icon=false}
## 🎯 Research Classification
- **Type**: Applied & Computational Research
- **Approach**: Quantitative Analysis
- **Design**: Quasi-Experimental with Time Series Validation
- **Validation**: Temporal Cross-Validation & Benchmarking
:::

---

## Research Framework

Our integrated methodology is structured in **six complementary stages**, each building upon previous findings to create a comprehensive portfolio optimization system.

### Stage 1: Data Collection & Preparation {#sec-data}

#### 1.1 Data Sources

We collect historical data from multiple authoritative sources:

- **Price Data**: B3 (Brasil Bolsa Balcão) futures contracts
- **Agricultural Indices**: CEPEA/ESALQ price indices
- **Macroeconomic Indicators**: USDA reports, Brazilian Central Bank data
- **Alternative Data**: News sentiment via APIs, weather patterns

#### 1.2 Commodities Universe

Our focus encompasses Brazilian agricultural grain commodities:

| Commodity | Contract Type | Data Frequency | Historical Depth |
|-----------|---------------|----------------|------------------|
| Corn | Futures (B3) | Daily | 15+ years |
| Soybeans | Futures (B3) | Daily | 15+ years |
| Soybean Meal | Futures (B3) | Daily | 10+ years |
| Soybean Oil | Futures (B3) | Daily | 10+ years |
| Coffee | Futures (B3) | Daily | 20+ years |
| Sugar | Futures (B3) | Daily | 15+ years |

#### 1.3 Data Preprocessing Pipeline


```{r}
#| eval: false
#| code-fold: show
#| code-summary: "Data Cleaning & Normalization Pipeline"

library(tidyverse)
library(quantmod)
library(xts)

# Data cleaning function
clean_commodity_data <- function(raw_data) {
  raw_data %>%
    # Remove outliers (> 5 SD)
    filter(abs(scale(returns)) < 5) %>%
    # Handle missing values via interpolation
    na.approx(maxgap = 5) %>%
    # Normalize to log-returns
    mutate(log_returns = log(close / lag(close))) %>%
    # Remove non-trading days
    filter(!is.na(log_returns))
}


```

::: {.callout-important}
## Data Quality Assurance
All datasets undergo rigorous quality checks including:
- Outlier detection via statistical thresholds
- Missing data imputation with maximum gap constraints
- Consistency validation across multiple sources
- Temporal alignment and synchronization
:::

---

### Stage 2: Time Series Forecasting Models {#sec-forecasting}

We employ a **hierarchical forecasting framework** combining econometric and machine learning approaches.

#### 2.1 Econometric Models

##### GARCH Family Models

For volatility forecasting, we implement multiple GARCH specifications:

**Standard GARCH(1,1)**:
$$
\sigma_t^2 = \omega + \alpha \epsilon_{t-1}^2 + \beta \sigma_{t-1}^2
$$

**GJR-GARCH** (capturing leverage effects):
$$
\sigma_t^2 = \omega + \alpha \epsilon_{t-1}^2 + \gamma \epsilon_{t-1}^2 I_{t-1} + \beta \sigma_{t-1}^2
$$

where $I_{t-1} = 1$ if $\epsilon_{t-1} < 0$, and 0 otherwise.

**Markov-Switching GARCH** (regime-dependent volatility):

```{r}
#| eval: false
#| code-fold: show

library(MSGARCH)

# MS-GARCH specification
spec <- CreateSpec(
  variance.spec = list(model = c("sGARCH", "sGARCH")),
  distribution.spec = list(distribution = c("std", "std")),
  switch.spec = list(K = 2)  # Two regimes
)

# Model estimation
fit <- FitML(spec, data = returns_data)

```

::: {.callout-tip}
## Why Markov-Switching?
Agricultural commodities exhibit **regime-dependent behavior** during crisis periods, supply shocks, or policy changes. MS-GARCH captures these structural breaks automatically.
:::

##### ARIMA Models with Exogenous Variables

For return forecasting, we use ARIMAX models incorporating:
- Lagged returns
- Macroeconomic indicators (USD/BRL exchange rate, interest rates)
- Seasonal components
- News sentiment scores

#### 2.2 Machine Learning Models

##### Long Short-Term Memory (LSTM) Networks

```{python}
#| eval: false
#| code-fold: show

import tensorflow as tf
from tensorflow.keras.models import Sequential
from tensorflow.keras.layers import LSTM, Dropout, Dense

def build_lstm_model(lookback=60, n_features=5):
    model = Sequential([
        LSTM(128, return_sequences=True, input_shape=(lookback, n_features)),
        Dropout(0.2),
        LSTM(64, return_sequences=False),
        Dropout(0.2),
        Dense(32, activation='relu'),
        Dense(1)  # Price/return prediction
    ])
    
    model.compile(optimizer='adam', loss='mse', metrics=['mae'])
    return model
  
```

##### Ensemble Methods

We combine multiple models via weighted averaging:

$$
\hat{y}_t = \sum_{i=1}^M w_i \hat{y}_{i,t}
$$

where weights $w_i$ are optimized via:
- **Inverse RMSE weighting**
- **Bayesian Model Averaging**
- **Stacking with meta-learner**

#### 2.3 Model Validation

::: {.panel-tabset}

## Walk-Forward Validation

- **Training Window**: Rolling 3-year window
- **Validation Period**: 6 months ahead
- **Re-estimation Frequency**: Monthly

## Performance Metrics

| Metric | Formula | Interpretation |
|--------|---------|----------------|
| RMSE | $\sqrt{\frac{1}{n}\sum(y_t - \hat{y}_t)^2}$ | Prediction error magnitude |
| MAE | $\frac{1}{n}\sum|y_t - \hat{y}_t|$ | Average absolute error |
| MAPE | $\frac{100}{n}\sum|\frac{y_t - \hat{y}_t}{y_t}|$ | Percentage error |
| Directional Accuracy | $\frac{1}{n}\sum I(sign(y_t) = sign(\hat{y}_t))$ | Correct direction % |

:::

---

### Stage 3: Multi-Objective Optimization Framework {#sec-optimization}

The portfolio optimization phase addresses **multiple conflicting objectives** simultaneously.

#### 3.1 Objective Functions

We optimize portfolios considering:

$$
\begin{aligned}
\text{Maximize:} \quad & f_1(w) = E[R_p] = w^T \mu \\
\text{Minimize:} \quad & f_2(w) = \text{CVaR}_\alpha(R_p) \\
\text{Minimize:} \quad & f_3(w) = \sigma_p = \sqrt{w^T \Sigma w} \\
\text{Maximize:} \quad & f_4(w) = \text{Diversification Ratio}
\end{aligned}
$$

Subject to:
$$
\begin{aligned}
\sum_{i=1}^n w_i &= 1 \\
0 \leq w_i &\leq w_{max} \\
\text{Turnover} &\leq \tau_{max}
\end{aligned}
$$

::: {.callout-note}
## Conditional Value-at-Risk (CVaR)
CVaR is preferred over VaR due to its:
- **Sub-additivity** (portfolio CVaR ≤ sum of individual CVaRs)
- **Convexity** (easier optimization)
- **Coherent risk measure properties**
:::

#### 3.2 Multi-Objective Algorithms

##### NSGA-II (Non-dominated Sorting Genetic Algorithm II)
```{r}
#| eval: false
#| code-fold: show

library(mco)

# Define multi-objective fitness function
fitness_function <- function(weights) {
  portfolio_return <- sum(weights * expected_returns)
  portfolio_risk <- sqrt(t(weights) %*% cov_matrix %*% weights)
  portfolio_cvar <- calculate_cvar(weights, returns_matrix, alpha = 0.05)
  
  return(c(
    -portfolio_return,  # Negative because we maximize
    portfolio_risk,
    portfolio_cvar
  ))
}

# Run NSGA-II
result <- nsga2(
  fn = fitness_function,
  idim = n_assets,  # Number of decision variables
  odim = 3,         # Number of objectives
  lower.bounds = rep(0, n_assets),
  upper.bounds = rep(0.4, n_assets),
  constraints = function(x) sum(x) - 1,  # Budget constraint
  popsize = 100,
  generations = 200
)

```

##### Differential Evolution for Multi-Objective Optimization

```{r}
#| eval: false

library(DEoptim)

# DE with Pareto ranking
de_result <- DEoptim(
  fn = fitness_function,
  lower = rep(0, n_assets),
  upper = rep(0.4, n_assets),
  control = DEoptim.control(
    strategy = 2,      # DE/rand/1/bin
    NP = 50,           # Population size
    itermax = 500,
    CR = 0.9,          # Crossover probability
    F = 0.8            # Mutation factor
  )
)
```

#### 3.3 Pareto Front Analysis

The optimization generates a **Pareto-optimal frontier** representing the trade-off between objectives:

```{r}
#| eval: false
#| code-fold: show

library(plotly)

# 3D Pareto Front visualization
plot_ly(
  data = pareto_solutions,
  x = ~expected_return,
  y = ~portfolio_risk,
  z = ~cvar,
  color = ~sharpe_ratio,
  type = "scatter3d",
  mode = "markers"
) %>%
  layout(
    title = "Pareto-Optimal Frontier",
    scene = list(
      xaxis = list(title = "Expected Return"),
      yaxis = list(title = "Risk (StdDev)"),
      zaxis = list(title = "CVaR (5%)")
    )
  )


```

---

### Stage 4: Reinforcement Learning for Dynamic Portfolio Management {#sec-rl}

We employ RL agents to learn **adaptive rebalancing strategies** that respond to market conditions.

#### 4.1 Markov Decision Process (MDP) Formulation

**State Space** $\mathcal{S}$:

$$
s_t = \{w_t, \mu_t, \Sigma_t, \text{indicators}_t\}
$$

**Action Space** $\mathcal{A}$:

$$
a_t = \Delta w_t \in [-\delta, \delta]^n \quad \text{(portfolio weight adjustments)}
$$

**Reward Function** $r_t$:

$$
r_t = R_{p,t} - \lambda \cdot \text{Risk}_{p,t} - \kappa \cdot \text{TransactionCost}_t
$$

#### 4.2 RL Algorithms

##### Deep Q-Network (DQN)

```{python}
#| eval: false
#| code-fold: show

import gym
import numpy as np
from stable_baselines3 import DQN
from stable_baselines3.common.vec_env import DummyVecEnv

# Custom environment for portfolio management
class PortfolioEnv(gym.Env):
    def __init__(self, price_data, initial_capital=100000):
        super(PortfolioEnv, self).__init__()
        
        self.price_data = price_data
        self.n_assets = price_data.shape[1]
        self.current_step = 0
        self.capital = initial_capital
        
        # Action: weight adjustments for each asset
        self.action_space = gym.spaces.Box(
            low=-0.1, high=0.1, shape=(self.n_assets,), dtype=np.float32
        )
        
        # State: prices, returns, portfolio weights, technical indicators
        self.observation_space = gym.spaces.Box(
            low=-np.inf, high=np.inf, 
            shape=(self.n_assets * 4,), 
            dtype=np.float32
        )
    
    def step(self, action):
        # Execute action, calculate reward, update state
        # ... implementation details ...
        return next_state, reward, done, info

# Train DQN agent
env = DummyVecEnv([lambda: PortfolioEnv(train_data)])
model = DQN("MlpPolicy", env, verbose=1, learning_rate=0.0001)
model.learn(total_timesteps=100000)

```

##### Proximal Policy Optimization (PPO)

```{python}
#| eval: false

from stable_baselines3 import PPO

# PPO for continuous action space
ppo_model = PPO(
    "MlpPolicy",
    env,
    learning_rate=0.0003,
    n_steps=2048,
    batch_size=64,
    n_epochs=10,
    gamma=0.99,
    gae_lambda=0.95,
    clip_range=0.2,
    verbose=1
)

ppo_model.learn(total_timesteps=200000)

```

#### 4.3 Strategy Types

The RL agent learns to implement various strategies:

| Strategy | Description | Trigger Conditions |
|----------|-------------|-------------------|
| **Dynamic Rebalancing** | Adjust weights based on forecasts | Deviation > threshold |
| **Momentum Trading** | Follow price trends | Strong directional signals |
| **Mean Reversion** | Contrarian positions | Extreme price movements |
| **Volatility Targeting** | Adjust exposure to volatility | Regime changes detected |
| **Hedging** | Risk mitigation positions | High uncertainty periods |

---

### Stage 5: Backtesting & Performance Evaluation {#sec-backtesting}

#### 5.1 Backtesting Protocol

::: {.callout-warning}
## Avoiding Overfitting
We implement strict protocols to prevent **backtest overfitting**:
- Out-of-sample testing with unseen data
- Transaction cost modeling (0.1% per trade)
- Realistic slippage assumptions
- No look-ahead bias in feature engineering
:::

```{r}
#| eval: false
#| code-fold: show

library(PerformanceAnalytics)

# Backtesting function
run_backtest <- function(strategy_weights, returns_data, costs = 0.001) {
  n_periods <- nrow(returns_data)
  portfolio_returns <- rep(0, n_periods)
  
  for (t in 2:n_periods) {
    # Calculate portfolio return
    portfolio_returns[t] <- sum(strategy_weights[t-1, ] * returns_data[t, ])
    
    # Subtract transaction costs
    turnover <- sum(abs(strategy_weights[t, ] - strategy_weights[t-1, ]))
    portfolio_returns[t] <- portfolio_returns[t] - (costs * turnover)
  }
  
  return(xts(portfolio_returns, order.by = index(returns_data)))
}

```

#### 5.2 Performance Metrics

##### Risk-Adjusted Returns

| Metric | Formula | Benchmark |
|--------|---------|-----------|
| **Sharpe Ratio** | $\frac{E[R_p - R_f]}{\sigma_p}$ | > 1.0 (good), > 2.0 (excellent) |
| **Sortino Ratio** | $\frac{E[R_p - R_f]}{\sigma_{downside}}$ | Higher is better |
| **Calmar Ratio** | $\frac{E[R_p]}{\text{Max Drawdown}}$ | > 0.5 (acceptable) |
| **Information Ratio** | $\frac{E[R_p - R_b]}{\text{TE}}$ | > 0.5 (outperformance) |

##### Drawdown Analysis

```{r}
#| eval: false

# Maximum drawdown calculation
calculate_drawdowns <- function(returns) {
  cumulative_returns <- cumprod(1 + returns)
  running_max <- cummax(cumulative_returns)
  drawdown <- (cumulative_returns - running_max) / running_max
  
  max_dd <- min(drawdown)
  max_dd_duration <- max(rle(drawdown < -0.05)$lengths)
  
  return(list(
    max_drawdown = max_dd,
    avg_drawdown = mean(drawdown[drawdown < 0]),
    max_duration = max_dd_duration
  ))
}

```

#### 5.3 Benchmark Comparisons

We compare our strategies against:

1. **Buy-and-Hold**: Equal-weighted static portfolio
2. **Mean-Variance (Markowitz)**: Traditional optimization
3. **Risk Parity**: Equal risk contribution
4. **1/N Portfolio**: Naïve diversification
5. **Momentum Strategy**: 12-month momentum signals

```{r}
#| eval: false
#| code-fold: show

library(ggplot2)

# Performance comparison plot
ggplot(performance_data, aes(x = date, y = cumulative_return, color = strategy)) +
  geom_line(size = 1.2) +
  scale_color_manual(values = c(
    "Our Strategy" = "#003d7a",
    "Buy-Hold" = "#6c757d",
    "Mean-Variance" = "#28a745",
    "Risk Parity" = "#ffc107"
  )) +
  labs(
    title = "Cumulative Returns: Strategy Comparison",
    x = "Date",
    y = "Cumulative Return",
    color = "Strategy"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    legend.position = "bottom"
  )

```

---

### Stage 6: Validation & Robustness Checks {#sec-validation}

#### 6.1 Sensitivity Analysis

Test model robustness under various scenarios:

- **Parameter Sensitivity**: Vary RL hyperparameters (±20%)
- **Window Length**: Test different training windows (2-5 years)
- **Transaction Costs**: Evaluate under 0%, 0.05%, 0.1%, 0.2% costs
- **Market Conditions**: Bull vs. bear vs. sideways markets

#### 6.2 Monte Carlo Simulation

```{r}
#| eval: false

# Monte Carlo portfolio simulation
monte_carlo_simulation <- function(n_sim = 10000, n_days = 252) {
  sim_results <- matrix(0, nrow = n_sim, ncol = n_days)
  
  for (i in 1:n_sim) {
    # Simulate returns based on estimated parameters
    sim_returns <- rmvnorm(n_days, mean = mu_hat, sigma = Sigma_hat)
    
    # Apply strategy
    sim_results[i, ] <- cumsum(apply_strategy(sim_returns))
  }
  
  # Calculate confidence intervals
  ci_lower <- apply(sim_results, 2, quantile, probs = 0.05)
  ci_upper <- apply(sim_results, 2, quantile, probs = 0.95)
  
  return(list(paths = sim_results, ci_lower = ci_lower, ci_upper = ci_upper))
}

```

#### 6.3 Out-of-Sample Validation

::: {.callout-tip}
## Cross-Validation Strategy
- **Training**: 2015-2020 (5 years)
- **Validation**: 2021-2022 (2 years)
- **Test (Out-of-Sample)**: 2023-2024 (2 years)

**Critical**: Test set is *never* used during model development.
:::

---

## Expected Outcomes

### Scientific Contributions

1. **Methodological Innovation**
   - Novel integration of forecasting + optimization + RL
   - Adaptive portfolio management framework
   - Regime-aware decision-making

2. **Empirical Insights**
   - Comprehensive analysis of Brazilian agricultural commodities
   - Benchmark comparisons across multiple strategies
   - Real-world applicability assessment

3. **Patent Development**
   - Unique algorithmic approach to multi-period optimization
   - Intellectual property documentation
   - Commercial application potential

### Academic Outputs

::: {.panel-tabset}

## Publications

**Target Journals:**

- European Journal of Operational Research

- Journal of Forecasting

- Quantitative Finance

- Agricultural Economics

- Journal of Commodity Markets

## Conferences

**Presentation Venues:**

- SPPAIC/FAE (Internal Symposium)

- Brazilian Finance Society (SBFin)

- Latin American Finance Network (LAFN)

- International Conference on Operational Research

## Technical Reports

- Quarterly progress reports

- Methodology documentation

- Reproducible code repositories

- Educational materials for students

:::

---

## Technical Stack & Tools

### Programming Languages

```{r}
#| eval: false
#| code-fold: show

# R Ecosystem
library(tidyverse)      # Data manipulation
library(quantmod)       # Financial data
library(rugarch)        # GARCH models
library(MSGARCH)        # Markov-Switching GARCH
library(PerformanceAnalytics)  # Portfolio analytics
library(DEoptim)        # Differential Evolution
library(mco)            # Multi-objective optimization

```


```{python}
#| eval: false
#| code-fold: show

# Python Ecosystem
import pandas as pd
import numpy as np
import tensorflow as tf
from stable_baselines3 import PPO, DQN
import pytorch
import gym
from sklearn.preprocessing import StandardScaler

```

### Development Infrastructure

- **Version Control**: GitHub (https://github.com/PAICEconometrics)
- **IDE**: RStudio / Positron IDE
- **Documentation**: Quarto Publishing System
- **Deployment**: GitHub Pages
- **Collaboration**: Slack + GitHub Projects

---

## Timeline & Milestones

```{mermaid}
gantt
    title PAIC Research Project Timeline (2025-2026)
    dateFormat  YYYY-MM-DD
    
    section Phase 1: Setup
    Data Collection & Infrastructure    :a1, 2025-08-01, 60d
    Literature Review                   :a2, 2025-08-01, 90d
    
    section Phase 2: Modeling
    Forecasting Models Development      :b1, 2025-10-01, 90d
    Multi-Objective Optimization        :b2, 2025-11-01, 120d
    Partial Report 1                    :milestone, m1, 2025-11-15, 1d
    
    section Phase 3: RL & Integration
    RL Environment Setup                :c1, 2026-02-01, 60d
    Agent Training & Ablation           :c2, 2026-02-15, 75d
    Partial Report 2                    :milestone, m2, 2026-03-15, 1d
    
    section Phase 4: Validation
    Backtesting & Robustness            :d1, 2026-03-01, 90d
    Seminar Presentation                :milestone, m3, 2026-04-15, 1d
    
    section Phase 5: Publication
    Article Writing                     :e1, 2026-04-01, 90d
    Patent Documentation                :e2, 2026-05-01, 60d
    Final Article Submission            :milestone, m4, 2026-07-15, 1d
```

---

## Quality Assurance

### Reproducibility Standards

All research outputs follow **FAIR principles** (Findable, Accessible, Interoperable, Reusable):

- ✅ **Code Repository**: Public GitHub with MIT License
- ✅ **Data Provenance**: Documented sources and timestamps
- ✅ **Environment Management**: `renv` / `conda` environment files
- ✅ **Literate Programming**: Quarto documents combining code + narrative
- ✅ **Version Control**: Semantic versioning for major milestones

### Ethical Considerations

::: {.callout-important}
## Research Ethics
- **No Market Manipulation**: Strategies are for academic purposes
- **Data Privacy**: Only publicly available market data
- **Transparency**: Full methodology disclosure
- **Conflict of Interest**: No undisclosed commercial relationships
:::

---

## Collaboration & Contact

Interested in collaborating or learning more about our methodology?

- **Principal Investigator**: Prof. Rodrigo Hermont Ozon
- **Email**: rodrigo.ozon@fae.edu
- **Institution**: FAE Business School, Curitiba, PR, Brazil
- **GitHub**: [@PAICEconometrics](https://github.com/PAICEconometrics)
- **LinkedIn**: [FAE Centro Universitário](https://www.linkedin.com/school/fae-centro-universitario/)

---

## References

::: {#refs}
:::
```{r}
#| echo: false
# This code block will render citations from references.bib
```

---

*Last updated: 

`r Sys.Date()`