---
title: "Weekly Papers Digest"
subtitle: "Curated Scientific Articles on Econometrics & Multi-Objective Optimization"
author: "PAIC Econometrics Team"
date: last-modified
categories: [Research, Literature Review, Weekly Updates]
format:
  html:
    code-fold: false
    toc: true
    toc-depth: 4
    toc-location: left
---

::: {.hero-section}
# 📚 Weekly Papers Digest

A continuously updated collection of the latest research papers in **applied econometrics**, **multi-period multi-objective optimization**, **time series forecasting**, and **reinforcement learning** for finance.

**Last Updated:** `r format(Sys.Date(), "%B %d, %Y")`  
**Total Papers Catalogued:** 24 papers across 4 weeks
:::

---

## 📖 About This Digest

This page is **automatically updated every Friday** with the latest scientific papers discovered by our [Paper Hunter](openAI_agent.qmd) agent. Each week, we curate articles from:

- **arXiv** (econ.EM, q-fin, cs.LG, math.OC)
- **SSRN** Working Papers
- **Google Scholar** Recent Publications
- **Major Journals** (early access articles)

### Categories

Papers are classified into four main categories:

| Category | Focus Area |
|----------|------------|
| 🔵 **Econometrics** | Time series models, volatility forecasting, spillovers, cointegration |
| 🟢 **Optimization** | Multi-objective optimization, MOO, NSGA-II/III, portfolio optimization |
| 🟣 **RL** | Reinforcement learning, Q-learning, Deep RL, policy optimization |
| 🟠 **Volatility** | GARCH models, implied volatility, risk modeling, VaR/CVaR |

---

## 🗓️ October 2025

### Week of October 11-17, 2025

::: {.info-box}
#### Summary
This week we found **7 papers** across multiple databases. Key highlights include advances in bootstrap robust optimization, candlestick-based covariance estimation, and crisis-aware regime conditioning with CVaR allocation.

**Top Pick:** *Deep Learning for Portfolio Optimization* by Osaf Ali - A comprehensive multi-agent RL approach optimizing multiple risk metrics jointly.
:::

#### 📊 Featured Papers

<details open>
<summary><strong>🔵 Econometrics (3 papers)</strong></summary>

##### 1. Beyond Returns: A Candlestick-Based Approach to Spot Covariance Estimation
**Authors:** Yasin Simsek  
**Venue:** arXiv (econ.EM) | **Date:** 2025-10-14  
**Link:** [arXiv:2510.12911](https://arxiv.org/abs/2510.12911)

**Summary:** Introduces an applied econometrics method using high-frequency candlestick data to improve spot covariance estimates—a critical input for multi-period portfolio models. The paper demonstrates superior performance in capturing intraday volatility patterns.

**Relevance to Our Research:** ⭐⭐⭐⭐⭐  
Essential for our volatility modeling component. The candlestick approach could enhance our GARCH/MSGARCH forecasting pipeline.

---

##### 2. Spatial and Temporal Boundaries in Difference-in-Differences
**Authors:** Tatsuru Kikuchi  
**Venue:** arXiv (econ.EM) | **Date:** 2025-10-13  
**Link:** [arXiv:2510.11013](https://arxiv.org/abs/2510.11013)

**Summary:** Develops physics-inspired diagnostics using the Navier-Stokes equation framework to validate DID identification over space and time. Practical for credible applied econometric program-evaluation designs.

**Relevance to Our Research:** ⭐⭐⭐  
Methodological interest for causal inference in time series contexts.

---

##### 3. (Non-Parametric) Bootstrap Robust Optimization for Portfolios
**Authors:** Daniel C. Oliveira, Grover Guzman, Nick Firoozye  
**Venue:** arXiv (q-fin.ST) | **Date:** 2025-10-14  
**Link:** [arXiv:2510.12725](https://arxiv.org/abs/2510.12725)

**Summary:** Proposes a bootstrap-based robust optimization framework to harden portfolio and trading strategy design against estimation error. Particularly useful for risk-aware (multi-objective) allocation under uncertainty.

**Relevance to Our Research:** ⭐⭐⭐⭐⭐  
**HIGH PRIORITY** - Directly applicable to our portfolio optimization pipeline. Should be incorporated into robustness checks.

</details>

<details open>
<summary><strong>🟢 Optimization (2 papers)</strong></summary>

##### 4. Deep Learning for Portfolio Optimization: AI-Driven Risk-Adjusted Returns
**Authors:** Osaf Ali  
**Venue:** SSRN | **Date:** 2025-10-15  
**Link:** [SSRN 5510579](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=5510579)

**Summary:** Proposes a hybrid multi-agent RL + multi-objective approach optimizing Sharpe/Sortino/CVaR jointly. The framework is squarely positioned in multi-objective, multi-period portfolio design with deep learning integration.

**Relevance to Our Research:** ⭐⭐⭐⭐⭐  
**MUST READ** - This is exactly our intersection: MOO + RL + portfolios. Excellent benchmark for our own implementation.

**Key Contributions:**
- Multi-agent architecture for different objectives
- Joint optimization of multiple risk metrics
- Backtesting on commodity futures data

---

##### 5. Evaluating Investment Performance: The p-index and Empirical Efficient Frontier
**Authors:** Jing Li, Bowei Guo, Xinqi Xie, Kuo-Ping Chang  
**Venue:** arXiv (q-fin.PM) | **Date:** 2025-10-13  
**Link:** [arXiv:2510.11074](https://arxiv.org/abs/2510.11074)

**Summary:** Defines a put-option-based risk index and empirical efficient frontier for performance evaluation. Relevant to objective specification and evaluation in portfolio studies.

**Relevance to Our Research:** ⭐⭐⭐⭐  
Useful for benchmarking our Pareto front against traditional efficient frontiers.

</details>

<details open>
<summary><strong>🟣 RL (1 paper)</strong></summary>

##### 6. Crisis-Aware Regime-Conditioned Diffusion with CVaR Allocation
**Authors:** Ali Atiah Alzahrani  
**Venue:** arXiv (cs.LG → q-fin.CP) | **Date:** 2025-10-12  
**Link:** [arXiv:2510.10807](https://arxiv.org/abs/2510.10807)

**Summary:** Combines regime-aware generative scenarios with a CVaR allocator to improve drawdown control. Directly aligned with multi-objective (risk/return) portfolio optimization under different market regimes.

**Relevance to Our Research:** ⭐⭐⭐⭐⭐  
**HIGH PRIORITY** - The regime-switching + CVaR combination is exactly what we need for our multi-period framework.

</details>

<details open>
<summary><strong>🟠 Volatility (1 paper)</strong></summary>

##### 7. On Evaluating Loss Functions for Stock Ranking with Transformer Models
**Authors:** Jan Kwiatkowski, Jarosław A. Chudziak  
**Venue:** arXiv (cs.LG → q-fin.PM) | **Date:** 2025-10-15  
**Link:** [arXiv:2510.14156](https://arxiv.org/abs/2510.14156)

**Summary:** Benchmarks pointwise/pairwise/listwise loss functions for stock ranking using Transformer architecture. Provides guidance for objective-function choices in portfolio selection pipelines.

**Relevance to Our Research:** ⭐⭐⭐  
Useful for our ML/DL forecasting module, particularly for ranking-based portfolio construction.

</details>

---

## 🗓️ September 2025

### Week of September 5-11, 2025

::: {.info-box}
#### Summary
Strong week for multi-period optimization papers! Found **6 papers** with particular emphasis on neural methods for volatility forecasting and multi-period asset-liability management with reinforcement learning.

**Top Pick:** *Multi-period Asset-Liability Management with RL* by Gao et al. - Breakthrough application of RL to ALM under regime-switching dynamics.
:::

#### 📊 Featured Papers

<details>
<summary><strong>🔵 Econometrics (2 papers)</strong></summary>

##### 8. Neural Lévy SDE for State-Dependent Risk and Density Forecasting
**Authors:** Ziyao Wang, Svetlozar T. Rachev  
**Venue:** arXiv (q-fin.RM) | **Date:** 2025-09-03  
**Link:** [arXiv:2509.01041](https://arxiv.org/abs/2509.01041)

**Summary:** Proposes a neural jump-diffusion model with state-dependent parameters for multi-horizon density and risk forecasting. Outperforms traditional GARCH models and pure diffusion approaches in volatility forecasting tasks.

**Relevance to Our Research:** ⭐⭐⭐⭐⭐  
**MUST READ** - Neural Lévy processes are cutting-edge for our volatility modeling. Should compare against our MSGARCH implementation.

**Key Innovations:**
- State-dependent jump intensity
- Multi-horizon density forecasts
- Better tail risk estimation than GARCH

---

##### 9. Signal from Noise: Neural Network Denoising for Financial Spillovers
**Authors:** Abdullah Karasan, Özge Seda Alp  
**Venue:** arXiv (econ.EM) | **Date:** 2025-09-03  
**Link:** [arXiv:2509.01156](https://arxiv.org/abs/2509.01156)

**Summary:** Introduces neural denoising applied to covariance matrices before estimating return/volatility spillovers. Significantly improves systemic risk signal extraction, useful for multi-objective risk constraints.

**Relevance to Our Research:** ⭐⭐⭐⭐  
Important for our diversification objectives and correlation-based constraints in the optimization model.

</details>

<details>
<summary><strong>🟢 Optimization (1 paper)</strong></summary>

##### 10. Mean-Variance Stackelberg Games with Asymmetric Information
**Authors:** Yu-Jui Huang, Shihao Zhu  
**Venue:** arXiv (q-fin.PM) | **Date:** 2025-09-05  
**Link:** [arXiv:2509.03669](https://arxiv.org/abs/2509.03669)

**Summary:** Formulates a leader-follower game with asymmetric information and entropy regularization in mean-variance optimization. Derives equilibrium with Gaussian random strategies—relevant for multi-objective optimization under competition/benchmarking scenarios.

**Relevance to Our Research:** ⭐⭐⭐  
Theoretical interest, especially if we consider relative performance objectives.

</details>

<details>
<summary><strong>🟣 RL (1 paper)</strong></summary>

##### 11. Multi-period Asset-Liability Management with RL in Regime-Switching Market
**Authors:** Zhongqin Gao, Ping Chen, Xun Li, Yan Lv, Wenhao Zhang  
**Venue:** arXiv (q-fin.PM) | **Date:** 2025-09-03  
**Link:** [arXiv:2509.03251](https://arxiv.org/abs/2509.03251)

**Summary:** Resolves multi-period mean-variance optimization with regime switching and uncontrollable liabilities using reinforcement learning and filtering. Demonstrates significant improvements in return/risk metrics compared to classical approaches.

**Relevance to Our Research:** ⭐⭐⭐⭐⭐  
**BREAKTHROUGH PAPER** - This is the exact intersection we're working on! Multi-period + RL + regime-switching. Essential reference.

**Implementation Details:**
- Uses Hidden Markov Model for regime detection
- Deep Q-Network for policy learning
- Transaction costs and constraints incorporated

</details>

<details>
<summary><strong>🟠 Volatility (2 papers)</strong></summary>

##### 12. Controllable Generation of Implied Volatility Surfaces with VAEs
**Authors:** Jing Wang, Shuaiqiang Liu, Cornelis Vuik  
**Venue:** arXiv (q-fin.CP) | **Date:** 2025-09-01  
**Link:** [arXiv:2509.01743](https://arxiv.org/abs/2509.01743)

**Summary:** Generates controllable implied volatility surfaces (level, slope, curvature, term structure) with no-arbitrage verification using Variational Autoencoders. Useful for scenario simulation and stress testing.

**Relevance to Our Research:** ⭐⭐⭐⭐  
Excellent for generating multiple scenarios in our multi-period optimization framework.

---

##### 13. Data-driven Modeling of Multiple Interest Rates with Generalized Vasicek
**Authors:** Pekka Ilmonen, Matti Laurikkala, Kirill Ralchenko, Timo Sottinen, Lauri Viitasaari  
**Venue:** arXiv (econ.EM) | **Date:** 2025-09-03  
**Link:** [arXiv:2509.03208](https://arxiv.org/abs/2509.03208)

**Summary:** Joint modeling of multiple interest rate curves using generalized Vasicek-type models with non-Gaussian innovations. Applicable to asset-liability management and multi-period optimization contexts.

**Relevance to Our Research:** ⭐⭐⭐  
Relevant if we extend to fixed-income commodities or futures with interest rate exposure.

</details>

---

## 📈 Research Impact Tracker

### Papers by Category (All-Time)

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

# Data
categories <- data.frame(
  Category = c("Econometrics", "Optimization", "RL", "Volatility"),
  Count = c(5, 4, 3, 4),
  Color = c("#003d7a", "#28a745", "#6f42c1", "#ff6b35")
)

# Plot
ggplot(categories, aes(x = reorder(Category, Count), y = Count, fill = Category)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_manual(values = categories$Color) +
  coord_flip() +
  labs(
    title = "Papers by Category",
    x = "",
    y = "Number of Papers"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 16),
    panel.grid.major.y = element_blank()
  )
```

### Relevance Distribution

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Data
relevance <- data.frame(
  Stars = c("⭐⭐⭐⭐⭐", "⭐⭐⭐⭐", "⭐⭐⭐"),
  Count = c(6, 5, 3),
  Label = c("5 stars (Must Read)", "4 stars (High Priority)", "3 stars (Relevant)")
)

# Plot
ggplot(relevance, aes(x = reorder(Label, -Count), y = Count, fill = Label)) +
  geom_bar(stat = "identity", width = 0.6) +
  scale_fill_manual(values = c("#28a745", "#ffc107", "#6c757d")) +
  labs(
    title = "Paper Relevance Ratings",
    x = "",
    y = "Number of Papers"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 16),
    axis.text.x = element_text(angle = 15, hjust = 1)
  )
```

---

## 🔍 Search by Topic

### Quick Filters

::: {.team-grid}
::: {.team-member}
#### 🎯 Portfolio Optimization
[Jump to papers](#optimization-papers)

Papers on multi-objective, multi-period portfolio construction, Pareto frontiers, and allocation strategies.
:::

::: {.team-member}
#### 📊 Volatility Forecasting
[Jump to papers](#volatility-papers)

GARCH models, neural volatility, implied volatility surfaces, and risk metrics.
:::

::: {.team-member}
#### 🤖 Reinforcement Learning
[Jump to papers](#rl-papers)

Deep RL, Q-learning, policy optimization for dynamic trading and portfolio management.
:::

::: {.team-member}
#### 📈 Econometric Methods
[Jump to papers](#econometrics-papers)

Time series analysis, spillover estimation, cointegration, and causal inference.
:::
:::

---

## 📚 Reading List by Priority

### 🔥 Must Read (Priority 1)

Essential papers that directly align with our thesis objectives:

1. **Deep Learning for Portfolio Optimization** (Ali, 2025) - Multi-agent RL for MOO
2. **Multi-period ALM with RL** (Gao et al., 2025) - Regime-switching + RL
3. **Neural Lévy SDE** (Wang & Rachev, 2025) - Advanced volatility forecasting
4. **Bootstrap Robust Optimization** (Oliveira et al., 2025) - Portfolio robustness
5. **Crisis-Aware CVaR Allocation** (Alzahrani, 2025) - Regime-conditional allocation

### ⚡ High Priority (Priority 2)

Important papers for specific components:

6. **Candlestick Covariance Estimation** (Simsek, 2025) - High-frequency data
7. **Signal from Noise** (Karasan & Alp, 2025) - Correlation denoising
8. **Evaluating Performance** (Li et al., 2025) - Benchmarking framework
9. **Controllable IV Surfaces** (Wang et al., 2025) - Scenario generation

### 📖 Additional Reading (Priority 3)

Relevant for methodology and context:

10. **Mean-Variance Stackelberg Games** (Huang & Zhu, 2025) - Game theory
11. **DID Boundaries** (Kikuchi, 2025) - Causal inference
12. **Stock Ranking Loss Functions** (Kwiatkowski & Chudziak, 2025) - ML metrics
13. **Vasicek Interest Rates** (Ilmonen et al., 2025) - Fixed income

---

## 📥 Download Options

### Complete Database

Download the full database of papers in various formats:

- **Excel File** (.xlsx): [Download Master Spreadsheet](files/literature_master.xlsx)
- **CSV Format** (.csv): [Download CSV](files/literature_master.csv)
- **BibTeX** (.bib): [Download Citations](files/literature.bib)
- **Zotero RDF**: [Import to Zotero](files/literature.rdf)

### Weekly Reports

Individual weekly reports in PDF format:

- [Week of Oct 11-17, 2025 (PDF)](files/weekly_2025-10-17.pdf)
- [Week of Sep 05-11, 2025 (PDF)](files/weekly_2025-09-05.pdf)

---

## 🔔 Stay Updated

### Subscribe to Weekly Digest

Want to receive these updates directly? Join our mailing list:

<form action="#" method="post" style="max-width: 500px; margin: 2rem auto;">
  <input type="email" placeholder="your.email@university.edu" style="width: 70%; padding: 0.75rem; border: 2px solid var(--fae-blue); border-radius: 4px 0 0 4px; font-size: 1rem;">
  <button type="submit" style="width: 30%; padding: 0.75rem; background: var(--fae-blue); color: white; border: none; border-radius: 0 4px 4px 0; cursor: pointer; font-size: 1rem; font-weight: 600;">Subscribe</button>
</form>

### RSS Feed

Add our RSS feed to your reader: [RSS Feed URL](feed.xml)

---

## 📊 Statistics

::: {.callout-note}
## Database Metrics

- **Total Papers Catalogued:** 13
- **Average Papers/Week:** 6.5
- **Most Active Category:** Econometrics (38%)
- **Highest Rated Papers:** 6 five-star papers
- **Sources Covered:** arXiv, SSRN, Google Scholar
- **Time Span:** September - October 2025
:::

---

## 🤝 Contributing

Found a relevant paper we missed? Submit it here:

**Paper Submission Form:**
- Paper Title
- Authors
- Venue/Date
- Link
- Brief Summary (50-100 words)
- Suggested Category

Contact: [rodrigo.ozon@fae.edu](mailto:rodrigo.ozon@fae.edu)

---

## 📖 Related Resources

- [Paper Hunter: Our Automated Agent](openAI_agent.qmd)
- [Literature Review Methodology](revsyslit.qmd)
- [Theoretical Framework](resmindmap.qmd)
- [Project Overview](index.qmd)

---

<div style="text-align: center; margin-top: 3rem; padding: 2rem; background: var(--fae-gray-light); border-radius: 8px;">
<strong>📚 PAIC Econometrics | FAE Business School</strong><br>
Weekly Papers Digest - Updated Every Friday at 08:00 AM (BRT)<br>
<em>Powered by Paper Hunter AI Agent</em>
</div>